---
title: "tmap basics: visual variables"
output: 
  bookdown::html_vignette2:
pkgdown:
  as_is: true
template:
  math-rendering: mathjax
bibliography: '`r system.file("tmap.bib", package="tmap")`'
csl: "`r system.file('ieee.csl', package = 'tmap')`"
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width=6, 
  fig.height=3,
  comment = "#>"
)
hook_output <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

```


```{r, echo = FALSE, message = FALSE}
library(tmap)
#devtools::load_all()
data(World, metro, rivers, land)
#tmap_design_mode()
```

## Introduction

### About the data

A spatial data object contained in tmap is called `World`. It is a data frame with a row for each country. The columns are the following data variables plus an additional geometry column which contains the geometries (see sf package):

```{r}
names(World)
```

We specify this object with `tm_shape` (see other vignette) and for convenience assign it to `s`:

```{r}
s = tm_shape(World, crs = "+proj=eqearth")
```


### Constant values

```{r}
s + tm_polygons(fill = "#ffce00", # fill color
				col = "black",    # line color
				lwd = 0.5,        # line width
				lty = "dashed")   # line type
```

For advanced users: the default constant values are specified for combinations of visual variables and layer type. See `tmap_options("value.const")`

## Visual variables

A visual variable describes a certain visual property of a drawn object, such as color, size, shape, line width, line stroke, transparency, fill pattern (in **ggplot2** these are called aesthetics). 
A visual variable can be specified using a constant value (e.g. \code{fill = "blue"}) or be **data-driven** (more on this later). If it can only be specified with a constant value, it is called a **visual constant**.

The following table shows which visual variables are used in standard map layers.



```{r}
s + tm_polygons(fill = "press")   # data variable (column in World)
```


```{r}
s + 
  tm_polygons(fill = "grey90") +   # fill color ()
  tm_bubbles(size = "pop_est",     # data variable, mapped to symbol size
  		     fill = "well_being",  # data variable, mapped to symbol fill color
  		     shape = "income_grp") # data variable, mapped to symbol shape

```

## Scales


## Facets

```{r, fig.height = 8}
s + tm_polygons(fill = c("well_being", "life_exp"))   # two data variables (columns in World)
```

<!-- (More on facets later) -->

## Legends


## Map modes 

## Base maps

## Layout

## Exporting maps
<!-- Plot and view -->


