---
title: "tmap: diff(v4, v3)"
author: "Martijn Tennekes"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    toc_depth: 2
link-citations: yes
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 9, warning = FALSE)
```

```{r, echo = FALSE, message = FALSE}
library(tmap)
#devtools::load_all()
data(World, metro, rivers, land)
#tmap_design_mode()
```


## tmap 4 - tmap 3.x

* **tmap 3.x** supports several map layer variables, for example `"col"`, `"size"`, and `"shape"` for the  `tm_symbols()` map layer function. There will be many more of those variables in **tmap 4**. Besides the visual variables, so-called transformation variables also will be available. A *transformation variable* role is to change the spatial coordinates (for instance, to create a cartogram). A *visual variable* only changes the appearance of a spatial object, e.g. fill color or line width. You can find some examples of these variables below.

* Map layer arguments (e.g. the arguments of `tm_polygons()`) are much better organized: for each visual/transformation variable, there are only four arguments. In case of the fill visual variable, these are: `fill`, `fill.scale`, `fill.legend` and `fill.free`, which respectively specify the data variable or visual value that defines the polygon fill color, the used scaling function, the legend layout, and whether scales are applied freely across facets.

* The input for each visual/transformation variable can be multivariate, in the sense that multiple data variables are scaled to one transformation of visual variable. An example is a bivariate choropleth, in which a cross tabulation of two data variables is mapped to one (bivariate) color palette.

* User-defined map layer functions can be written (e.g. as an extension package).

* Like **tmap 3.x**, **tmap 4.0** comes with two modes, a `"plot"` and a `"view"` mode. However, other modes can be added as well, so you may expect an extension package **tmap.rayshader** at some point (or start writing one yourself).

* Legends and other map components (such as scale bars) can be drawn anywhere on or outside the map.

* The layout of legends has been improved and made much more flexible.

* It is possible to combine legends, which is useful if the same data variable is applied for multiple visual variables using the same scaling function.

