---
title: "tmap: diff(v4, v3)"
output: 
  bookdown::html_vignette2:
pkgdown:
  as_is: true
template:
  math-rendering: mathjax
bibliography: '`r system.file("tmap.bib", package="tmap")`'
csl: "`r system.file('ieee.csl', package = 'tmap')`"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width=6, 
  fig.height=3,
  comment = "#>"
)
hook_output <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

```

```{r, echo = FALSE, message = FALSE}
library(tmap)
#devtools::load_all()
data(World, metro, rivers, land)
#tmap_design_mode()
```

## tmap 4 - tmap 3.x

-   **tmap 3.x** supports several map layer variables, for example `"col"`, `"size"`, and `"shape"` for the `tm_symbols()` map layer function. 
There will be many more of those variables in **tmap 4**. 
Besides the visual variables, so-called transformation variables also will be available.
A *transformation variable* role is to change the spatial coordinates (for instance, to create a cartogram). 
A *visual variable* only changes the appearance of a spatial object, e.g. fill color or line width.
You can find some examples of these variables below.

-   Map layer arguments (e.g. the arguments of `tm_polygons()`) are much better organized: for each visual/transformation variable, there are only four arguments.
In the case of the fill visual variable, these are: `fill`, `fill.scale`, `fill.legend` and `fill.free`, which respectively specify the data variable or visual value that defines the polygon fill color, the used scaling function, the legend layout, and whether scales are applied freely across facets.

-   The input for each visual/transformation variable can be multivariate, in the sense that multiple data variables are scaled to one transformation of a visual variable.
An example is a bivariate choropleth, in which a cross tabulation of two data variables is mapped to one (bivariate) color palette.

-   User-defined map layer functions can be written (e.g. as an extension package).

-   Like **tmap 3.x**, **tmap 4.0** comes with two modes, a `"plot"` and a `"view"` mode.
However, other modes can be added as well, so you may expect an extension package **tmap.rayshader** at some point (or start writing one yourself).

-   Legends and other map components (such as scale bars) can be drawn anywhere on or outside the map.

-   The layout of legends has been improved and made much more flexible.

-   It is possible to combine legends, which is useful if the same data variable is applied to multiple visual variables using the same scaling function.
