---
title: "tmap basics: legends"
output: 
  bookdown::html_vignette2:
pkgdown:
  as_is: true
template:
  math-rendering: mathjax
bibliography: '`r system.file("tmap.bib", package="tmap")`'
csl: "`r system.file('ieee.csl', package = 'tmap')`"
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  out.width = "100%",
  dpi = 300,
  fig.width = 7.2916667,
  comment = "#>"
)
hook_output <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

```


```{r, echo = FALSE, message = FALSE}
library(tmap)
tmap_options(scale = 0.75)
```



### About the data

A spatial data object contained in tmap is called `World`. It is a data frame with a row for each country. The columns are the following data variables plus an additional geometry column which contains the geometries (see sf package):

```{r}
names(World)
```

We specify this object with `tm_shape` (see other vignette) and for convenience assign it to `s`:

```{r}
s = tm_shape(World, crs = "+proj=eqearth")
```


## Legend

Each visual variable, e.g. `fill` in `tm_polygons`, also has a `.legend` argument. The most popular properties are illustrated in the next example: 

```{r, fig.height = 3.5}
s + tm_polygons(
  fill = "HPI",
  fill.scale = tm_scale_continuous(values = "pu_gn"),
  fill.legend = 
    tm_legend(
      title = "Happy Planex Index", 
      orientation = "landscape", 
      width = 60))
```

The legend width is set in number of text line heights, which corresponds to the number of characters. In other words, the resulting width depends on the fontsize, which can be set in `tm_legend`, but also via the general `scale` option, which determines the overall 'scale' of the map (all line widths, point sizes and font sizes):

```{r, fig.height = 3.5}
s + tm_polygons(
  fill = "HPI",
  fill.scale = tm_scale_continuous(values = "pu_gn"),
  fill.legend = 
    tm_legend(
      title = "Happy Planex Index", 
      orientation = "landscape", 
      width = 60)) +
tm_layout(scale = 0.25)
```

## Combine legends

Suppose we want to use both `fill` and `shape` to show `"economy"`. Then we can combine the legends as follows:

```{r, fig.height = 3.5}
s + 
  tm_polygons() +
  tm_symbols(
    fill = "economy",
    shape = "economy",
    shape.legend = tm_legend_combine("fill"))
```

## Custom visual values

Suppose, for whatever reason, we would like to change some of the visual values used in the legend symbols. For instance, instead of solid circle lines, thick dotted lines. In that case, we can simply pass on the corresponding visual values `lwd` and `lty` to `tm_legend()`:

```{r, fig.height = 3.5}
s + 
  tm_polygons() +
  tm_symbols(
    fill = "economy",
    fill.legend = tm_legend("Economy", lwd = 2, lty = "dotted"))
```

## Manual legends

```{r, fig.height = 3.5}
s + 
  tm_polygons() +
  tm_add_legend(title = "Countries", fill = "grey80", shape = 22, size = c(1, 2, 4), labels = c("Small", "Medium", "Large"))
```
