---
title: "tmap: scales"
author: "Martijn Tennekes"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    toc_depth: 2
link-citations: yes
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 9, warning = FALSE)
```

```{r, echo = FALSE, message = FALSE}
library(tmap)
#devtools::load_all()
data(World, metro, rivers, land)
#tmap_design_mode()
```


## Scales

Each visual variable and each transformation variable can be scaled with one of the `tm_scale_` functions. To illustrate the different options, we show life expectancy across Africa, which we round in order to use the categorical scales as well.

```{r, message=FALSE}
data(World)
Africa = World[World$continent == "Africa", ]
Africa$life_exp = round(Africa$life_exp)
```

Like **tmap 3.x**, it is possible to create facets by specifying multiple data variable names and scales to one visual (or transformation) variable, in this case `"fill"`:

```{r, class.source='bg-success',message=FALSE,fig.height=7}
tm_shape(Africa) +
	tm_polygons(rep("life_exp", 6), 
				fill.scale = list(tm_scale_categorical(),
								  tm_scale_ordinal(),
								  tm_scale_intervals(),
								  tm_scale_continuous(),
								  tm_scale_continuous_log(),
								  tm_scale_discrete()),
				fill.legend = tm_legend(title = "", position = tm_pos_in("left", "top"))) +
	tm_layout(panel.labels = c("tm_scale_categorical", "tm_scale_ordinal", "tm_scale_intervals", "tm_scale_continuous", "tm_scale_continuous_log", "tm_scale_discrete"), 
			  inner.margins = c(0.05, 0.4, 0.1, 0.05),
			  legend.text.size = 0.5)
```

Both `tm_scale_categorical()` and `tm_scale_ordinal()` tread data as categorical data, so ignoring the fact that they are actually numbers. The only difference is that categorical does not assume any order between the categories, whereas ordinal does. This is similar to a factor in R which can be ordered or not.

The other shown scales can only be applied to numeric data. Note that in this example the breaks of `tm_scale_intervals()` are similar to the tick marks of `tm_scale_continous()`. However, when using class intervals only a few colors are used (in this case 6 plus a color for missing values) whereas in a continuous scale a gradient of colors is used. The advantage of using class intervals is that it is relatively easy to read data values from the map, e.g. the value of South Africa is 55 to 60, while the advantage of using a continuous color scale is that the colors in the map are more accurate (because they are unrounded).

For `tm_scale_intervals()` it is possible to chose how to determine the breaks (with the argument `style`). For `tm_scale_continous()` it is possible to use a transformation function: in this case the built-in log transformation is used (which is pretty useless for this particular example because of the data range).

Finally, `tm_scale_discrete()` uses a discrete linear scale. Note that this is different than `tm_scale_ordinal()`, which does not use colors for values that are not present (as categories), for instance 53.

Each `tm_scale_*()` functions can (in principle) be applied to any visual or transformation variable. Note that this is different from **ggplot2** where scales are organized by variable and by type (e.g. `ggplot2::scale_fill_continuous()`). This is related to another difference with **ggplot2**. In **tmap**, the scales are set directly in the map layer function to the target visual/transformation variable, for instance `tm_polygons(fill = "x", fill.scale = tm_scale_continuous())`. In `ggplot()`, scales are set outside the layer functions.

Each `tm_scale_` function has (at least) the following arguments: `values`, `values.repeat`, `values.range`, `values.scale`, `value.na`, `value.null`, `value.neutral`, `labels`, `label.na`, `label.null`, and `label.format`.
The `value*` arguments determine the visual values to which the data values are mapped. In case the scale is applied to a visual variable that represents color, they takes color values or a color palette. However, if for instance the same scale is applied to line width, then values should be numeric values that represent line widths.

This is illustrated in the following example:

```{r}
tm_shape(World) +
	tm_polygons(fill = "HPI", fill.scale = tm_scale_intervals(values = "scico.roma", value.na = "grey95", breaks = c(12,20,30,45))) +
	tm_symbols(size = "HPI", size.scale = tm_scale_intervals(values = c(0.3,0.5, 0.8), value.na = 0.1, breaks = c(12,20,30,45)), col = "grey30")
```

The defaults for those `value.*` arguments are stored in the **tmap** options. For instance

```{r}
tmap_options("values.var")$values.var$fill
```

contains the default color palettes for the visual variable `"fill"` for different types of data. For instance, when data values are all positive numbers, and `tm_scale_intervals()` or `tm_scale_continuous()` is applied, the default color palette is `"hcl.blues3"`, as can be seen in the examples above.

Regarding the available color palettes: **tmap** uses the new R package **cols4all** which contains a large number of well-known color palettes. 
Please run `cols4all::c4a_gui()` which starts an interactive tool (the successor of `tmaptools::palette_explorer()`). 
Of course, also own color palettes can be loaded directly via a vector of color codes.
