---
title: "tmap: facets"
author: "Martijn Tennekes"
output: 
  bookdown::html_vignette2:
pkgdown:
  as_is: true
template:
  math-rendering: mathjax
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width=6, 
  fig.height=3,
  comment = "#>",
  fig.path = "man/figures"
)
hook_output <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

```


```{r, echo = FALSE, message = FALSE}
library(tmap)
#devtools::load_all()
data(World, metro, rivers, land)
#tmap_design_mode()
```


```{r echo=FALSE,message=FALSE,fig.height = 2}
World$pop_class = cut(World$pop_est, breaks = c(0, 10, 100, 1000, Inf) * 1e6, labels = c("Small", "Medium", "Large", "Extra Large"))					   
World$HPI_class = cut(World$HPI, breaks = seq(10, 50, by = 10))
World$well_being_class = cut(World$well_being, breaks = seq(2, 8, by = 2))
World$footprint_class = cut(World$footprint, breaks = seq(0, 16, by = 4))

metro$pop2020_class = cut(metro$pop2020, breaks = c(.5, 1.5, 2.5, 5, 15, 40) * 1e6)



Africa = World[World$continent == "Africa", ]
```

# Facetting shapes

These tests focus on dealing with shapes when facets are created.
Legends, scales will be studied in other tests.

## Polygons

### One facet

```{r}
tm_shape(World) +
	tm_polygons("HPI")
```


### Multiple facets, specified with multiple variables

```{r}
# one polygon shape, two aes values
tm_shape(World) +
	tm_polygons(c("HPI", "income_grp"))
```



### Multiple facets, specified by a 'facet wrap'

```{r}
# split polygon shape (by)
tm_shape(World) +
	tm_polygons("HPI") +
	tm_facets(by = "continent")
```

### Multiple facets, specified by a 'facet wrap', with fixed coordinates

```{r}
# split polygon shape (by), with fixed coords
tm_shape(World) +
	tm_polygons("HPI") +
	tm_facets(by = "continent", free.coords = FALSE)
```




### Filter



```{r}
# filter polygons (filtered out will get value.null)
tm_shape(World, filter = World$pop_est > 100e6) +
	tm_polygons(c("blue", "red"))
```


```{r}
tm_shape(World, filter = World$pop_est > 100e6) +
	tm_polygons("HPI")
```

```{r}
tm_shape(World, filter = World$pop_est > 100e6) +
	tm_polygons(c("HPI", "economy"))
```


### Drop units


```{r}
# fixed aes, wrap
tm_shape(World) +
	tm_polygons("blue") +
	tm_facets(by = "continent", free.coords = FALSE, drop.units = FALSE)
```

```{r}
# one aesthetics, different frees
tm_shape(World) +
	tm_polygons("HPI") +
	tm_facets(by = "continent", free.coords = FALSE, drop.units = FALSE)
```


```{r}
# multiple aesthetics, different frees
tm_shape(World) +
	tm_polygons("HPI", lty = "economy", fill.free = FALSE, lty.free = TRUE) +
	tm_facets(by = "continent", free.coords = FALSE, drop.units = FALSE)
```


```{r}
# nothing to drop
tm_shape(World) +
	tm_polygons(c("HPI", "economy")) +
	tm_facets(free.coords = FALSE, drop.units = TRUE)
```


### Drop empty facets
```{r}
# default drop empty (wrap) facets
tm_shape(Africa) +
	tm_polygons("blue") +
	tm_facets(by = "income_grp")

tm_shape(Africa) +
	tm_polygons("HPI") +
	tm_facets(by = "income_grp")
```

```{r}
# keep empty (wrap) facets
tm_shape(Africa) +
	tm_polygons("blue") +
	tm_facets(by = "income_grp", drop.empty.facets = FALSE)

tm_shape(Africa) +
	tm_polygons("HPI") +
	tm_facets(by = "income_grp", drop.empty.facets = FALSE)
```

```{r}
# default drop empty (xtab) facet: only if whole row or column is empty
tm_shape(Africa) +
	tm_polygons("pop_est_dens") +
	tm_facets_grid("income_grp", "well_being_class", drop.units = FALSE)
```

```{r}
# keep empty (xtab) facets
tm_shape(Africa) +
	tm_polygons("pop_est_dens") +
	tm_facets_grid("income_grp", "well_being_class", drop.units = FALSE, drop.empty.facets = FALSE)
```


### Drop NA facets

```{r}
# default keep NA facets (where by variable is NA)
tm_shape(Africa) +
	tm_polygons("blue") +
	tm_facets(by = "footprint_class")
```


```{r}
# drop NA facets 
tm_shape(Africa) +
	tm_polygons("blue") +
	tm_facets(by = "footprint_class", drop.NA.facets = TRUE)
```

```{r}
# default keep NA facets (where by variable is NA)
tm_shape(Africa) +
	tm_polygons("pop_est_dens") +
	tm_facets(by = "footprint_class")
```


```{r}
# drop NA facets 
tm_shape(Africa) +
	tm_polygons("pop_est_dens") +
	tm_facets(by = "footprint_class", drop.NA.facets = TRUE)
```


```{r}
# default keep NA facets (where by variable is NA)
tm_shape(Africa) +
	tm_polygons("pop_est_dens") +
	tm_facets_grid("footprint_class", "HPI_class")
```


```{r}
# drop NA facets 
tm_shape(Africa) +
	tm_polygons("pop_est_dens") +
	tm_facets_grid("footprint_class", "HPI_class", drop.NA.facets = TRUE)
```





## Rasters

```{r}
# nothing to drop
tm_shape(land) +
	tm_raster("cover_cls")
```
