---
title: "tmap: exporting maps"
output: 
  bookdown::html_vignette2:
pkgdown:
  as_is: true
template:
  math-rendering: mathjax
bibliography: '`r system.file("tmap.bib", package="tmap")`'
csl: "`r system.file('ieee.csl', package = 'tmap')`"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width=6, 
  fig.height=3,
  comment = "#>"
)
hook_output <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

```

```{r, echo = FALSE, message = FALSE}
library(tmap)
#devtools::load_all()
data(World, metro, rivers, land)
#tmap_design_mode()
```

With the **tmap** package, thematic maps can be generated with great flexibility.
The syntax for creating plots is similar to that of **ggplot2**, but tailored to maps.
This vignette is for those who want to get started with **tmap** within a couple of minutes.

For more context on R's geographic capabilities we recommend the online version of the book [Geocomputation with R](https://r.geocompx.org/). 
The [Making maps with R](https://r.geocompx.org/adv-map) chapter of the book provides many more context and abundant code examples of map making with **tmap** and other packages. 
Other good resources are the vignettes of the [**sf** package](https://CRAN.R-project.org/package=sf/vignettes/sf1.html).

### Hello World!

A good place to start is to create a map of the world.
After [installing](https://github.com/r-tmap/tmap#installation) **tmap**, the following lines of code should create the map shown below:

```{r}
library(tmap)
tm_shape(World) +
    tm_polygons("HPI")
```

The object `World` is a spatial object of class `sf` from the [**sf** package](https://CRAN.R-project.org/package=sf); it is a `data.frame` with a special column that contains a geometry for each row, in this case polygons. 
In order to plot it in **tmap**, you first need to specify it with `tm_shape()`.
Layers can be added with the `+` operator, in this case, `tm_polygons().`
There are many layer functions in **tmap**, which can easily be found in the documentation by their `tm_` prefix.
See also `?'tmap-element'`.

### Interactive maps

Each map can be plotted as a static image or viewed interactively using `"plot"` and `"view"` modes, respectively.
The mode can be set with the function `tmap_mode`, and toggling between the modes can be done with the 'switch' `ttm()` (which stands for toggle thematic map).

```{r}
tmap_mode("view")

tm_shape(World) +
    tm_polygons("HPI")
```

### Multiple shapes and layers

A shape is a spatial object (with a class from `sf`, `stars`, or `terra`).
Multiple shapes and also multiple layers per shape can be plotted:

```{r}

tmap_mode("plot")
tm_shape(land) +
    tm_raster("elevation", palette = terrain.colors(10)) +
tm_shape(World) +
    tm_borders("white", lwd = .5) +
    tm_text("iso_a3", size = "AREA") +
tm_shape(metro) +
    tm_symbols(col = "red", size = "pop2020", scale = .5) +
tm_legend(show = FALSE)
```

### Facets

Facets can be created in three ways:

1. By assigning multiple variable names to one aesthetic (in this example the first argument of `tm_polygons()`:

```{r}
tmap_mode("view")
tm_shape(World) +
    tm_polygons(c("HPI", "economy")) +
    tm_facets(sync = TRUE, ncol = 2)
```

2. By splitting the spatial data with the `by` argument of `tm_facets`: 

```{r}
tmap_mode("plot")

tm_shape(NLD_dist) +
    tm_polygons("pop_65plus", palette = "RdYlBu") +
    tm_facets(by = "province")
```

3. By using the `tmap_arrange` function:

```{r}
tmap_mode("plot")

tm1 <- tm_shape(NLD_muni) + tm_polygons("population", convert2density = TRUE)
tm2 <- tm_shape(NLD_muni) + tm_bubbles(size = "population")

tmap_arrange(tm1, tm2)
```

### Basemaps and overlay tile maps

Tiled basemaps can be added with the layer function `tm_basemap`.
Semi-transparent overlay maps (for example annotation labels) can be added with `tm_tiles`.

```{r}
tmap_mode("view")
tm_basemap("Stadia.StamenWatercolor") +
tm_shape(metro) + tm_bubbles(size = "pop2020", col = "red") +
tm_tiles("Stadia.StamenTonerLabels")
```


See a [preview of the available tilemaps](https://leaflet-extras.github.io/leaflet-providers/preview/). 
This list is also accessible in R: `leaflet::providers`.

### Options and styles

The functions `tm_layout()` and `tm_view()` are used to specify the map layout and the interactive aspects respectively.
These functions can be used in the same way as the layer functions, e.g.

```{r}
tmap_mode("plot")

tm_shape(World) +
    tm_polygons("HPI") +
tm_layout(bg.color = "skyblue", inner.margins = c(0, .02, .02, .02))
```

These options, as well as a couple of others, can also be set within with `tmap_options`, which works in the same way as the base R function `options`.
The main advantage is that these options are set globally, so they do not have to be specified in each map, for the duration of the session.

```{r}
tmap_options(bg.color = "black", legend.text.color = "white")

tm_shape(World) +
    tm_polygons("HPI", legend.title = "Happy Planet Index")
```

A style is a certain configuration of the tmap options. 

```{r}
tmap_style("classic")

tm_shape(World) +
    tm_polygons("HPI", legend.title = "Happy Planet Index")
```

```{r}
# see what options have been changed
tmap_options_diff()

# reset the options to the default values
tmap_options_reset()
```

New styles can be created; see `?tmap_options`.

### Exporting maps

```{r}
tm <- tm_shape(World) +
    tm_polygons("HPI", legend.title = "Happy Planet Index")

tm

ttmp()
if (FALSE) {
  ## save an image ("plot" mode)
  tmap_save(tm, filename = "world_map.png")

## save as stand-alone HTML file ("view" mode)
tmap_save(tm, filename = "world_map.html")
}

```

### Shiny integration

It is possible to use tmap in shiny:

```{r, eval = FALSE}
# in UI part:
tmapOutput("my_tmap")

# in server part
output$my_tmap = renderTmap({
    tm_shape(World) + tm_polygons("HPI", legend.title = "Happy Planet Index")
})
```

See `?renderTmap` for a working example.

### Quick thematic map

Maps can also be made with one function call.
This function is `qtm`:

```{r}
qtm(World, fill = "HPI", fill.palette = "RdYlGn")
```

### Tips 'n Tricks

Run:

```{r}
tmap_tip()
```
