---
title: "tmap example: choropleth"
output: 
  bookdown::html_vignette2:
pkgdown:
  as_is: true
template:
  math-rendering: mathjax
bibliography: '`r system.file("tmap.bib", package="tmap")`'
csl: "`r system.file('ieee.csl', package = 'tmap')`"
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width=6, 
  fig.height=3,
  comment = "#>"
)
hook_output <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

```

```{r, message = FALSE}
library(tmap)
library(dplyr)
library(sf)
tmap_options(scale = 0.5)
```


### About the data

In tmap, there are two example data sets of the Netherlands:

* one for districts, `NLD_dist` (3340 in total),
* one for municipalities: `NLD_muni` (345 in total), and 
* one for provinces: `NLD_prov` (12 in total).

The first two datasets contain demographic data of the Netherlands as of 2022. The third dataset only contains metadata (names and codes) and will only be used to plot province borders. 

```{r, out.width="100%"}
tmap_arrange(
	qtm(NLD_dist, title = "NLD_dist: 3340 districts"),
	qtm(NLD_muni, title = "NLD_mini: 345 municipalities"),
	qtm(NLD_prov, title = "NLD_prov: 12 provinces"),
	ncol = 3
)
```


The variables in the `NLD_dist` and `NLD_muni` are the same:

```{r}
names(NLD_dist)
```

We will create a chroropleth about the variable `"edu_appl_sci"`, which contains the percentage of 15-75 year old people who have a university degree (Dutch: WO) or applied sciences degree (Dutch: HBO), per district as of 1st October 2022.

## The choropleth

```{r, fig.height = 8, out.width="100%"}
tm_shape(NLD_dist) +
    tm_polygons(
        fill = "edu_appl_sci",
        fill.scale = tm_scale_continuous(values = "-orange_blue_diverging"),
        fill.legend = tm_legend("", position = tm_pos_in("left", "top")),
        col = NULL # disable border lines
    ) + 
tm_shape(NLD_muni) +
	tm_borders(lwd = 1) +
tm_shape(NLD_prov) +
	tm_borders(lwd = 3) + 
tm_title("Proportion of Population Aged 15–75 with a University or Applied Sciences Degree by District (as of October 1, 2022)", width = 15, position = tm_pos_in("left", "top"), z = 0) +
tm_compass(position = tm_pos_in("left", "bottom")) +
tm_scalebar(position = tm_pos_in("left", "bottom")) +
tm_credits("© Data: Statistics Netherlands, Software: R-tmap", position = tm_pos_in("left", "bottom"))
```
