---
title: "tmap foundations: Grammar of Graphics"
output: 
  bookdown::html_vignette2:
pkgdown:
  as_is: true
template:
  math-rendering: mathjax
bibliography: '`r system.file("tmap.bib", package="tmap")`'
csl: "`r system.file('ieee.csl', package = 'tmap')`"
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width=6, 
  fig.height=3,
  comment = "#>"
)
hook_output <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

```


```{r, echo = FALSE, message = FALSE}
library(tmap)
#devtools::load_all()
data(World, metro, World_rivers, land)
#tmap_design_mode()
```

```{r, echo = FALSE}
plot_tmap_grammar = function(asp = 5/7, scale = 0.7) {
	library(grid)
	grid.newpage()
	
	if (asp < 1) {
		pushViewport(viewport(width = unit(asp, "snpc"), height = unit(1, "npc"), gp = gpar(cex = scale)))
	} else {
		pushViewport(viewport(width = unit(1, "snpc"), height = unit(1/asp, "npc"), gp = gpar(cex = scale)))
	}
	
	#grid.rect()
	
	cellplot = function (r, c, e, ...) {
	    grid::pushViewport(grid::viewport(layout.pos.row = r, layout.pos.col = c, 
	        clip = TRUE, ...))
	    e
	    grid::upViewport()
	}
	
	
	hs = local({
		tmp = rep(c(2,1), 14)
		tmp / sum(tmp)
	})
	
	ws = local({
		tmp = c(0.5, 4, 0.5, 
				1, 5, 
				1, 3)
		tmp / sum(tmp)
	})
	
	pushViewport(viewport(layout = grid.layout(ncol = length(ws), nrow = length(hs), widths = ws, heights = hs)))
	
	text_block = function(r, c, text, x = 0.2) {
		cellplot(r, c, {
			grid.rect(gp = gpar(fill = "grey90"))
			grid.text(text, x = x, just = "left")
		})
	}
	text_block_reg_it = function(r, c, text_reg, text_it, xmid = 0.3) {
		cellplot(r, c, {
			grid.rect(gp = gpar(fill = "grey90"))
			grid.text(text_reg, x = xmid, just = "right")
			grid.text(text_it, x = xmid, just = "left", gp = gpar(fontface = "italic"))
		})
	}
	text_only = function(r, c, text) {
		cellplot(r, c, {
			grid.text(text)
		})
	}
	text_only_from_left = function(r, c, text, x = 0.2) {
		cellplot(r, c, {
			grid.text(text, x = x, just = "left")
		})
	}
	
	text_only_list = function(r, c, texts) {
		n = length(texts)
		cellplot(r, c, {
			pushViewport(viewport(layout = grid.layout(nrow = n+2, heights = unit(c(0.5, rep(1.5, n), 0.5), units = c("null", rep("lines", n), "null")))))
			for (i in 1:n) {
				cellplot(i+1, 1, {
					grid.text(texts[i], x = 0.2, just = "left")		
				})
			}
			upViewport()
		})
	}
	brackets = function(r, c) {
		m = 0.05
		cellplot(r, c, {
			grid.lines(c(0.3, 0.6, 0.6, 0.3), c(m, m, 1-m, 1-m), gp = gpar(lwd = 2))
			grid.lines(c(0.6, 0.8), c(0.5, 0.5), gp = gpar(lwd = 2))
		})
	}
	
	col_block1 = 1:2
	col_block2 = 2:3
	
	text_block(1, col_block1, "tm_shape")
	text_block_reg_it(3, col_block2, "tm_", "layer_1", xmid = 0.2)
	text_block_reg_it(5, col_block2, "tm_", "layer_2", xmid = 0.2)
	text_only(6, col_block2, "...")
	text_block_reg_it(7, col_block2, "tm_", "layer_k", xmid = 0.2)
	text_block(9, col_block2, "tm_facets")
	text_block_reg_it(11, col_block1, "tm_", "layer_aux_1")
	text_block_reg_it(13, col_block1, "tm_", "layer_aux_2")
	text_only(14, col_block1, "...")
	text_block_reg_it(15, col_block1, "tm_", "layer_aux_k")
	
	text_block_reg_it(17, col_block1, "tm_", "component_1")
	text_block_reg_it(19, col_block1, "tm_", "component_2")
	text_only(20, col_block1, "...")
	text_block_reg_it(21, col_block1, "tm_", "component_k")
	text_block(23, col_block1, "tm_style")
	text_block(25, col_block1, "tm_layout")
	#text_block(25, col_block1, "tm_options")
	#text_block(25, col_block1, "tm_view")
	
	
	
	
	text_only_list(1, 7, "tm_shape")
	text_only_list(3:7, 7, c("tm_polygons", "tm_symbols", "tm_lines", "tm_text", "tm_raster", "..."))
	text_only_list(9, 7, c("tm_facets_wrap", "tm_facets_grid"))
	text_only_list(11:15, 7, c("tm_basemap", "tm_tiles", "tm_grid", "tm_graticules", "..."))
	text_only_list(17:21, 7, c("tm_compass", "tm_scalebar", "tm_credits", "tm_logo", "..."))
	text_only_list(23, 7, c("tm_style"))
	text_only_list(25, 7, c("tm_layout", "tm_view", "tm_plot", "tm_options"))
	
	text_only_from_left(1, 5, "Data and coordinates")
	brackets(3:7, 4)
	text_only_from_left(3:7, 5, "Data-driven map layers")
	brackets(11:15, 4)
	text_only_from_left(9, 5, "Facets (small multiples)")
	text_only_from_left(11:15, 5, "Auxiliary map layers")
	brackets(17:21, 4)
	text_only_from_left(17:21, 5, "Map components")
	text_only_from_left(23, 5, "Overall style")
	text_only_from_left(25, 5, "Layout options")
}
```

The Grammar of Graphics [@wilkinson2005] is a framework for building statistical graphics by combining key components like data, aesthetic mappings, geometric objects, scales, transformations, and statistical summaries. It provides a systematic, modular approach to creating flexible and consistent visualizations.

### Components:
- **Data**: The dataset being visualized.
- **Aesthetic Mappings**: How data maps to visual properties (e.g., position, color, size).
- **Geometric Objects**: Shapes representing the data (e.g., points, lines, bars).
- **Scales**: Rules linking data values to aesthetic values.
- **Transformations**: Adjustments to data or coordinates (e.g., log scales).
- **Statistical Summaries**: Computed data representations (e.g., means, trends).
- **Facets**: Layouts for splitting data into subsets for comparison.

It is applied in **ggplot2** [@ggplot2] and also in **tmap**, but slightly differently.

## tmap

```{r echo=FALSE, fig.width = 7, fig.height=7}
plot_tmap_grammar(asp = 7/7, scale = .8)
```

## ggplot2

* In ggplot2, the visual variables are defined on plot level (by default), but in tmap on layer level. This makes sense, since in non-spatial visualizations the x and y variables, the most important ones, are shared among the plot layers. However, in spatial visualizations the x and y variables are not considered data-driven but rather geometry-driven. The other visual variables, such as fill and border color, line width, and symbol shape.
* This also applies to the scale funtions: e.g. in ggplot2, the `scale_fill_continuous()` is defined for the visual variable fill for the whole plot. In tmap, the scale functions are mapped 1 to 1 to the visual variables per layer: `tm_polygons(fill = "my_var", fill.scale = tm_scale_continuous())`

See other vignette (to do: link) that compares tmap with ggplot2 by a series of examples

