---
title: "tmap versus: ggplot2"
output: 
  bookdown::html_vignette2:
pkgdown:
  as_is: true
template:
  math-rendering: mathjax
bibliography: '`r system.file("tmap.bib", package="tmap")`'
csl: "`r system.file('ieee.csl', package = 'tmap')`"
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  out.width = "100%",
  dpi = 300,
  fig.width = 7.2916667,
  comment = "#>"
)
hook_output <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

```


```{r, echo = FALSE, message = FALSE}
library(tmap)
tmap_options(scale = 1)
```

## Choropleth Map comparision

An example of how the default output of `tmap` compares with `ggplot2`. 

This is the default output of `ggplot2`:

```{r}
library(ggplot2)
library(cols4all)

ggplot(World) +
	geom_sf(aes(fill = HPI), color = "grey20", linewidth = .4) +
	scale_fill_continuous_c4a_div("pu_gn", mid = 35) +
	coord_sf(crs = "+proj=eqearth") +
	theme_void()
```


This is the default output of `tmap`:

```{r}
tm_shape(World, crs = "+proj=eqearth") +
	tm_polygons(fill = "HPI",
				col = "grey20",
				lwd = 1,
				fill.scale = tm_scale_continuous(values = "pu_gn", midpoint = 35))
```

Note the different line width values: 

- `linewidth = .4` in `ggplot2`
- `lwd = 1` in `tmap`

The unit of a line width is different. `tmap` follows the `lwd` parameter (see [`graphics::par`](https://www.rdocumentation.org/packages/graphics/versions/3.6.2/topics/par)) whereas in `ggplot2` `linewidth = 1` equals roughly 0.75 [due to a historical error](https://ggplot2.tidyverse.org/articles/ggplot2-specs.html#linewidth). 


## Mimicking ggplot2 layout

We can use `tmap` to match the style of `ggplot2`:

```{r}
tm_shape(World, crs = "+proj=eqearth") +
	tm_polygons(fill = "HPI",
				col = "grey20",
				lwd = 1,
				fill.scale = tm_scale_continuous(values = "pu_gn", midpoint = 35),
				fill.legend = tm_legend(reverse = TRUE, 
										frame = FALSE, 
										item.height = 2.25, 
										item.width = 1.8, 
										position = tm_pos_out(pos.v = "center"), 
										na.show = FALSE, 
										ticks = list(c(0, 0.1), c(0.9, 1)), 
										ticks.col = "white", 
										col = "white")) +
	tm_layout(frame = FALSE, outer.margins = 0)
```


