---
title: "tmap versus: mapview"
output: 
  bookdown::html_vignette2:
pkgdown:
  as_is: true
template:
  math-rendering: mathjax
bibliography: '`r system.file("tmap.bib", package="tmap")`'
csl: "`r system.file('ieee.csl', package = 'tmap')`"
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  out.width = "100%",
  dpi = 300,
  fig.width = 7.2916667,
  comment = "#>"
)
hook_output <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

```


```{r, echo = FALSE, message = FALSE}
library(tmap)
tmap_options(scale = 1)
```

## Mapview

Mapview is an excellent R package for interactive maps. Although the packages have a lot in common, the focus is different:

* tmap focusses on thematic mapping with a syntax that is based on the [grammar of graphics](21_theory_gg) and is therefore popular for education purposes;
* mapview focusses on fast exploration of spatial data and has more interactive features.

## Modes / platforms

* tmap offers two modes: "plot" and "view" (but is extendable, see https://github.com/r-tmap/tmap.deckgl). 
* mapview supports three modes, which they call *platforms*: "leaflet", "leafgl", and "mapdeck".

tmap "view" (with `tm_view(use_WebGL = FALSE)`) is similar to mapview "leaflet"
tmap "view" (with `tm_view(use_WebGL = TRUE)`) is similar to mapview "leafgl"

tmap does not offer a mode using Mapbox yet.

## Default maps

This is the default output of `mapview`:

```{r eval=FALSE}
library(mapview)
mapview(World)
```

```{r fig.height = 3.5, eval=TRUE, echo=FALSE}
library(mapview)
lf = mapview(World)@map
lf
```




This is the default output of `tmap`:

```{r eval=FALSE}
tmap_mode("view")
qtm(World) # qmt stands for 'quick thematic map'
```


```{r fig.height = 3.5, eval=TRUE, echo=FALSE}
tmap_mode("view")
lf = tmap_leaflet(qtm(World))
lf
```

## Choropleth

```{r eval=FALSE}
mapview(World, zcol = "HPI")
```

```{r fig.height = 3.5, eval=TRUE, echo=FALSE}
lf = mapview(World, zcol = "HPI")@map
lf
```

```{r eval=FALSE}
tm_shape(World) +
  tm_polygons(fill = "HPI")
```

```{r fig.height = 3.5, eval=TRUE, echo=FALSE}
lf = tmap_leaflet(tm_shape(World) + tm_polygons(fill = "HPI"))
lf
```

## Mimicking mapview layout

We can use `tmap` to match the style of `mapview`:

```{r eval=FALSE}
tm_shape(World) +
  tm_polygons(
    fill = "HPI",
    fill_alpha = 0.6,
    col_alpha = 0.9,
    fill.legend = tm_legend(
      title = "World - HPI",
      position = c("right", "top"), 
      fill_alpha = 1),
    fill.scale = tm_scale_continuous(values = "viridis", n = 7, value.na = "#BEBEBE")
  ) + 
tm_basemap(c("CartoDB.Positron", "CartoDB.DarkMatter", 
  "OpenStreetMap", "Esri.WorldImagery", "OpenTopoMap"))
```

```{r fig.height = 3.5, eval=TRUE, echo=FALSE}
lf = tmap_leaflet(tm_shape(World) +
  tm_polygons(
    fill = "HPI",
    fill_alpha = 0.6,
    col_alpha = 0.9,
    fill.legend = tm_legend(
      title = "World - HPI",
      position = c("right", "top"), 
      fill_alpha = 1),
    fill.scale = tm_scale_continuous(values = "viridis", n = 7, value.na = "#BEBEBE")
  ) + 
tm_basemap(c("CartoDB.Positron", "CartoDB.DarkMatter", 
  "OpenStreetMap", "Esri.WorldImagery", "OpenTopoMap")))
lf
```
