---
title: "tmap advanced: layer blending"
output: 
  bookdown::html_vignette2:
pkgdown:
  as_is: true
template:
  math-rendering: mathjax
bibliography: '`r system.file("tmap.bib", package="tmap")`'
csl: "`r system.file('ieee.csl', package = 'tmap')`"
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  out.width = "100%",
  dpi = 300,
  fig.width = 7.2916667,
  comment = "#>"
)
hook_output <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

```


```{r, echo = FALSE, message = FALSE}
library(tmap)
tmap_options(scale = 0.75)
```

## What is blending?

By default, when one map layer is drawn on top of another, the top layer simply covers the layer below. Where the top layer is (partly) transparent, the two are combined with *alpha compositing*: the result is a weighted average based on the opacity.

**Blending** generalizes this. Instead of averaging, a blend mode defines a *formula* that combines the color of each pixel in the top layer with the color of the pixel beneath it, producing a single combined image. This is the same idea as the "blend modes" in image editors such as GIMP or Photoshop, and as the layer compositing in the [ggblend](https://mjskay.github.io/ggblend/) package for ggplot2.

Blending is set with the `blend` argument, which is available on every layer function (`tm_polygons()`, `tm_raster()`, `tm_rgb()`, `tm_symbols()`, `tm_lines()`, `tm_text()`, etc. [#815](https://github.com/r-tmap/tmap/issues/815)).

## Blend modes

tmap supports the blend modes below. The "Result" column describes what the combined image looks like when the top layer is blended onto the layer below. The default `"over"` applies no blending, so existing maps are unaffected.

| `blend` | Result |
| --- | --- |
| `"over"` | The top layer simply covers the layer below (default; no blending). |
| `"multiply"` | Always darker. White in the top layer leaves the layer below unchanged; black turns it black. |
| `"screen"` | Always lighter. The opposite of multiply. |
| `"overlay"` | Raises contrast: dark areas get darker, light areas get lighter. |
| `"darken"` | Keeps the darker color at each pixel. |
| `"lighten"` | Keeps the lighter color at each pixel. |
| `"color.dodge"` | Strongly brightens the layer below. |
| `"color.burn"` | Strongly darkens the layer below. |
| `"hard.light"` | Strong contrast, driven by the top layer. |
| `"soft.light"` | Soft contrast; a subtle version of `hard.light`. |
| `"difference"` | The absolute difference of the two layers. Identical colors become black. |
| `"exclusion"` | Like difference but with lower contrast. |

The mode names above are the ones used in **plot mode**. In **view mode** the compound names follow the CSS spelling with hyphens instead of dots (`"color-dodge"`, `"color-burn"`, `"hard-light"`, `"soft-light"`); the others are identical.

The exact formula behind each mode is given in the [appendix](#appendix-blend-mode-formulas).

## A sandbox example

The clearest way to see what each mode does is to blend two rasters where we know exactly what the pixels are. We use:

* a **color image** (the bottom layer): a vivid rainbow that runs through every hue from left to right, and
* a **white-to-black gradient** (the top layer): white at the top, black at the bottom.

Both are built with **terra** on the same grid, so they align pixel for pixel:

```{r, message = FALSE}
library(terra)

n = 300

# bottom layer: a horizontal rainbow (hue varies with x, full saturation/brightness)
img = rast(nrows = n, ncols = n, xmin = 0, xmax = 1, ymin = 0, ymax = 1, nlyrs = 3)
xy = xyFromCell(img, 1:ncell(img))
cols = grDevices::hsv(h = xy[, 1], s = 1, v = 1)
values(img) = t(grDevices::col2rgb(cols))
names(img) = c("red", "green", "blue")

# top layer: a vertical white-to-black gradient
grad = rast(img, nlyrs = 1)
values(grad) = rep(seq(1, 0, length.out = nrow(grad)), each = ncol(grad))
names(grad) = "gradient"
```

We map the gradient with a continuous black-to-white scale, so a value of `1` is white (top) and `0` is black (bottom). Here are the two layers on their own:

```{r, fig.height = 4}
tm_rainbow = tm_shape(img) +
	tm_rgb()	

tm_gradient = tm_shape(grad) +
  tm_raster("gradient",
    col.scale = tm_scale_continuous(values = c("black", "white")),
    col.legend = tm_legend_hide())

tmap_arrange(tm_rainbow, tm_gradient)
```

Now we blend the gradient onto the rainbow. Here are all twelve modes side by side; the first panel (`over`) is the default, where the gradient simply covers the rainbow. A small helper keeps the code compact:

```{r, fig.height = 5.5}
blend_map = function(mode) {
  tm_shape(img) +
    tm_rgb() +
  tm_shape(grad) +
    tm_raster("gradient",
      col.scale = tm_scale_continuous(values = c("black", "white")),
      col.legend = tm_legend_hide(),
      blend = mode) +
  tm_title(mode, size = 0.8) +
  tm_layout(frame = FALSE)
}

modes = c("over", "multiply", "screen", "overlay",
          "darken", "lighten", "color.dodge", "color.burn",
          "hard.light", "soft.light", "difference", "exclusion")

tmap_arrange(lapply(modes, blend_map), ncol = 4)
```

Because the gradient is white at the top and black at the bottom, each panel reads top-to-bottom:

* **`multiply`** keeps the rainbow at the top (multiplying by white = 1) and fades it to black at the bottom (multiplying by black = 0).
* **`screen`** does the reverse: untouched at the bottom, washed out to white at the top.
* **`overlay`** and **`hard.light`** push contrast, darkening the lower half and brightening the upper half.
* **`darken`** and **`lighten`** keep, at each pixel, whichever of the two layers is darker or lighter.
* **`color.dodge`** and **`color.burn`** are the strong versions of brightening and darkening.
* **`difference`** and **`exclusion`** subtract the two layers, so the colors invert where the gradient is bright.

## Example 1: basemap labels

A very common reason to reach for `multiply` is to drape a thematic layer over a basemap while keeping the basemap's reference detail — place names, roads, country borders — readable underneath. Light basemaps such as `CartoDB.Positron` draw that detail as dark marks on a near-white background. An opaque thematic fill would hide it completely, and even a semi-transparent fill dulls it.

`multiply` solves this neatly. Multiplying the thematic color by the near-white background barely changes it, so the choropleth shows in full where the basemap is empty. Multiplying it by the dark labels and lines keeps those dark, so they stay legible right through the fill.

We map the Happy Planet Index of the `World` dataset over Africa. First the basemap and the choropleth on their own:

```{r, fig.height = 4}
# the basemap on its own
tm_basemap("CartoDB.Positron", zoom = 3) +
	tm_crs(bbox = "Africa", ext = 2)
```

```{r, fig.height = 4}
# the choropleth on its own
tm_shape(World, bbox = "Africa", ext = 2) + 
	tm_polygons(fill = "HPI", 
				fill.scale = tm_scale_intervals(values = "-bu_br_div"))
```

Now the choropleth blended onto the basemap with `multiply`. The labels and borders from the basemap remain visible through the colors:

```{r, fig.height = 4}
# multiply overlay
tm_basemap("CartoDB.Positron", zoom = 3) +
	tm_shape(World, bbox = "Africa", ext = 2) + 
	tm_polygons(fill = "HPI", 
				fill.scale = tm_scale_intervals(values = "-bu_br_div"),
				blend = "multiply")
```

## Example 2: shaded relief

A common reason to reach for `blend` is to combine a colored map with a *hillshade*: a grayscale layer that simulates light and shadow on terrain, making slopes visible. On its own the hillshade is just gray; blended onto a colored elevation map it adds a sense of relief without changing the colors.

We compute the hillshade from the elevation raster in the `land` dataset:

```{r, message = FALSE}
elev = rast(land)["elevation"]
elev[is.na(elev)] = 0

hs = shade(
  slope  = terrain(elev, "slope",  unit = "radians"),
  aspect = terrain(elev, "aspect", unit = "radians"))
names(hs) = "hillshade"
```

First, the two layers on their own. The colored elevation map (left) and the grayscale hillshade (right):

```{r, fig.height = 3}
m_elev = tm_shape(land) +
  tm_raster("elevation",
    col.scale = tm_scale_continuous(values = "hcl.terrain"),
    col.legend = tm_legend_hide()) +
  tm_title("elevation")

m_hs = tm_shape(hs) +
  tm_raster("hillshade",
    col.scale = tm_scale_continuous(values = c("black", "white")),
    col.legend = tm_legend_hide()) +
  tm_title("hillshade")

tmap_arrange(m_elev, m_hs, ncol = 2)
```

Now we draw the elevation map and blend the hillshade on top with `"overlay"`, which darkens the shaded slopes and lightens the lit ones while keeping the terrain colors:

```{r, fig.height = 4.5}
tm_shape(land) +
  tm_raster("elevation",
    col.scale = tm_scale_continuous(values = "hcl.terrain"),
    col.legend = tm_legend()) +
tm_shape(hs) +
  tm_raster("hillshade",
    col.scale = tm_scale_continuous(values = c("black", "white")),
    col.legend = tm_legend_hide(),
    blend = "overlay") +
tm_crs("+proj=eqearth") +
tm_layout(
  earth_boundary = TRUE,
  bg.color = "slategray1",
  frame = FALSE,
  earth_boundary.lwd = 1, 
  inner.margins = 0.01)
```

## Requirements and notes

Blending in plot mode relies on graphics-device *compositing*, which has a few requirements:

* **R >= 4.2** is needed. On older versions `blend` is ignored (with a warning) and the layer is drawn normally.
* A **compatible graphics device** is required, for example `png(type = "cairo")` or `svg()`. If the active device does not support the requested operator, tmap falls back to no blending and warns. The default RStudio graphics device may not support compositing; if a blended map looks unblended, re-plot to a Cairo or SVG device. The simpler operators (`"multiply"`, `"screen"`, ...) are the most widely supported; the compound ones (`"color.dodge"`, `"soft.light"`, ...) require a fuller compositing implementation.
* Blending acts on the *rendered pixels*, so it is most predictable when the layers involved are opaque. Combining `blend` with partial transparency (`fill_alpha`, `col_alpha`) is possible but harder to reason about.

Internally, plot mode uses [`grid::groupGrob()`][grid::groupGrob], which also supports the Porter-Duff operators (`"clear"`, `"source"`, `"in"`, `"out"`, `"atop"`, `"xor"`, ...) beyond the blend modes listed above.

### View mode

In interactive (`"view"`) mode, blending is applied through the CSS `mix-blend-mode` property on the Leaflet pane of the layer. The simple mode names are identical to plot mode; the compound ones use the CSS hyphen spelling (`"color-dodge"`, `"hard-light"`, ...). As in plot mode, `"over"` means no blending.

```{r, eval = FALSE}
tmap_mode("view")

tm_shape(land) +
  tm_raster("elevation",
    col.scale = tm_scale_continuous(values = "hcl.terrain")) +
tm_shape(hs) +
  tm_raster("hillshade",
    col.scale = tm_scale_continuous(values = c("black", "white")),
    col.legend = tm_legend_hide(),
    blend = "overlay")
```

## Appendix: blend mode formulas

For each pixel, let $S$ be the source (the top layer) and $D$ the destination (the layer below it), both as RGB values normalized to $[0, 1]$; $\alpha$ is the opacity of the top layer.

| `blend` | Formula |
| --- | --- |
| `"over"` | $S \cdot \alpha + D \cdot (1 - \alpha)$ |
| `"multiply"` | $S \cdot D$ |
| `"screen"` | $1 - (1 - S)(1 - D)$ |
| `"overlay"` | multiply if $D < 0.5$, screen otherwise |
| `"darken"` | $\min(S, D)$ |
| `"lighten"` | $\max(S, D)$ |
| `"color.dodge"` | $D / (1 - S)$ |
| `"color.burn"` | $1 - (1 - D) / S$ |
| `"hard.light"` | overlay with $S$ and $D$ swapped |
| `"soft.light"` | gentle `hard.light` |
| `"difference"` | $\lvert S - D \rvert$ |
| `"exclusion"` | $S + D - 2 S D$ |
