---
title: "tmap advanced: positioning of components"
output: 
  bookdown::html_vignette2:
pkgdown:
  as_is: true
template:
  math-rendering: mathjax
bibliography: '`r system.file("tmap.bib", package="tmap")`'
csl: "`r system.file('ieee.csl', package = 'tmap')`"
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  out.width = "100%",
  dpi = 300,
  fig.width = 7.2916667,
  comment = "#>"
)
hook_output <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

```


```{r, echo = FALSE, message = FALSE}
library(tmap)
tmap_options(scale = 0.75)
```

## Component positionining

Recall from the [vignette about components](https://r-tmap.github.io/tmap/articles/07_basic_components) that the `position` argument of map components and legends is used to position them.

Let's enable the 'design mode' of tmap for the rest of this vignette

```{r}
tmap_design_mode()
```


## Shortcut: vector of two


A vector of two, where the first is the horizontal position and the second the vertical position. These shortcut will place the component inside the map frame. They are a shortcut for `tm_pos_in`, explained below


### Character

(`"left"`, `"center"`, `"right"`), the second one the vertical position (`"top"`, `"center"`, `"bottom"`)

```{r, fig.height = 6}
tm_shape(NLD_muni) +
  tm_polygons(
    fill = "edu_appl_sci",
    fill.legend = tm_legend(position = c("left", "top"))) +
tm_credits("Statistics Netherlands (CBS)", position = c("left", "bottom")) +
tm_compass(type = "8star", position = c("right", "bottom")) +
tm_scalebar(position = c("right", "bottom"))
```

### Numeric

It is also possible to specify two numbers to set the location. The coordinates are between 0 and 1, where (0,0) is bottom left.

```{r, fig.height = 6}
tm_shape(NLD_muni) +
  tm_polygons(
    fill = "edu_appl_sci",
    fill.legend = tm_legend(position = c(0.05, 0.7))) +
tm_credits("Statistics Netherlands (CBS)", position = c(0.05, 0.05)) +
tm_compass(type = "8star", position = c(0.8, 0.2)) +
tm_scalebar(position = c(0.3, 0.15))
```


## Advanced positioning with `tm_pos()`

`tm_pos_in()` draws the component inside the map frame and `tm_pos_out()` draws the component outside the map frame.

#### `cell.h` and `cell.v`

The total plot area (i.e. device) is a 3 x 3 grid. The map is draw in the middle grid cell.

```{r, fig.height = 4}
tm1 = tm_shape(NLD_muni) + tm_polygons() +
		tm_compass(type = "8star", size = 4, position = tm_pos_out(cell.h = "left", cell.v = "center"))
tm2 = tm_shape(NLD_muni) + tm_polygons() +
		tm_compass(type = "8star", size = 4, position = tm_pos_out(cell.h = "center", cell.v = "bottom"))
tm3 = tm_shape(NLD_muni) + tm_polygons() +
		tm_compass(type = "8star", size = 4, position = tm_pos_out(cell.h = "right", cell.v = "bottom"))
tmap_arrange(tm1, tm2, tm3, ncol = 3)
```

Notes: 

* the sizes of the cells are automatically determined based on the content and the margins (see [vignette](https://r-tmap.github.io/tmap/articles/42_margins))
* `tm_pos_in()` is a shortcut for `tm_pos_out()` with both `cell.h` and `cell.v` set to `"center"`.

#### `pos.h` and `pos.v`

The position of a component within a cell:

```{r, fig.height = 4}
tm_shape(NLD_muni) + tm_polygons() +
		tm_compass(type = "8star", size = 8, position = tm_pos_out(cell.h = "left", cell.v = "center", pos.v = "bottom"))
```

The values can either be characters or numbers (as explained [above](https://r-tmap.github.io/tmap/articles/41_adv_positions#shortcut-vector-of-two)).

### `just.h` and `just.v`

In case `pos.h` and `pos.v` are numbers, the justification can be set with `just.h` and `just.v`:

```{r, fig.height = 4}
tm1 = tm_shape(NLD_muni) + tm_polygons() +
		tm_compass(type = "8star", size = 4, position = tm_pos_in(pos.h = .5, pos.v = .5, just.h = "center", just.v = "center"))
tm2 = tm_shape(NLD_muni) + tm_polygons() +
		tm_compass(type = "8star", size = 4, position = tm_pos_in(pos.h = .5, pos.v = .5, just.h = "left", just.v = "top"))
tm3 = tm_shape(NLD_muni) + tm_polygons() +
		tm_compass(type = "8star", size = 4, position = tm_pos_in(pos.h = .5, pos.v = .5, just.h = "right", just.v = "bottom"))
tmap_arrange(tm1, tm2, tm3, ncol = 3)
```


### `align.h` and `align.v`

The alignment of components within the same cell:

```{r, fig.height = 4}
tm1 = tm_shape(NLD_muni) + tm_polygons() +
	tm_compass(type = "8star", position = tm_pos_in(pos.h = "right", pos.v = "bottom", align.h = "left")) +
	tm_scalebar(position = tm_pos_in(pos.h = "right", pos.v = "bottom", align.h = "left"))
tm2 = tm_shape(NLD_muni) + tm_polygons() +
	tm_compass(type = "8star", position = tm_pos_in(pos.h = "right", pos.v = "bottom", align.h = "center")) +
	tm_scalebar(position = tm_pos_in(pos.h = "right", pos.v = "bottom", align.h = "left"))
tm3 = tm_shape(NLD_muni) + tm_polygons() +
	tm_compass(type = "8star", position = tm_pos_in(pos.h = "right", pos.v = "bottom", align.h = "right")) +
	tm_scalebar(position = tm_pos_in(pos.h = "right", pos.v = "bottom", align.h = "left"))
tmap_arrange(tm1, tm2, tm3, ncol = 3)
```


## Automatic positioning

The default position of legends and map components are in some cases done automatically. Automatic positioning can be achieved by setting a `.position` argument to `tm_pos_auto_in()` or `tm_pos_auto_out()`. The former checks in which map corner is most space available (and bases `pos.h` and `pos.v` on that). The latter calculates in which grid cell (so `cell.h` and `cell.v`) the map components can be positioned, which is based on the aspect ratio and margins.
