---
title: "tmap advanced: ggplot2 integration"
output: 
  bookdown::html_vignette2:
pkgdown:
  as_is: true
template:
  math-rendering: mathjax
bibliography: '`r system.file("tmap.bib", package="tmap")`'
csl: "`r system.file('ieee.csl', package = 'tmap')`"
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  out.width = "100%",
  dpi = 300,
  fig.width = 7.2916667,
  comment = "#>"
)
hook_output <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

```

```{r, echo = FALSE, message = FALSE}
library(tmap)
tmap_options(scale = 0.75)
```

Recall from the [comparison](https://r-tmap.github.io/tmap/articles/versus_ggplot2) that tmap and ggplot are both flexible packages. The ggplot2 layout can be [mimicked by tmap](https://r-tmap.github.io/tmap/articles/versus_ggplot2#mimicking-ggplot2-layout) and probably the other way round as well. 

In this vignette we'll show how to use **ggplot2** inside **tmap**.

## tmap charts

ggplot2 is already used by tmap to draw [charts](https://r-tmap.github.io/tmap/articles/basics_charts). Via the argument `extra.ggplot2` additional ggplot2 code can be provided:

```{r, fig.height = 6}
library(ggplot2)

tm_shape(NLD_muni) +
  tm_polygons("edu_appl_sci",
    fill.legend = tm_legend("University degree"),
    fill.chart = tm_chart_bar(extra.ggplot2 = theme(panel.grid.major.y = element_line(colour = "red")))) 
```


## ggplot2 charts

Much more flexibility is achieved when adding a ggplot2 plot via `tm_inset()`. First, let's create a violin plot of the distribution of higher education across municipalities, grouped by level of urbanity. 

```{r, fig.height = 3}
(g1 = ggplot(NLD_muni, aes(x = urbanity, y = edu_appl_sci, fill = urbanity)) +
	geom_violin() +
	scale_x_discrete(labels = c("++", "+", "o", "-", "--")) +
 	scale_fill_brewer(guide = "none", type = "div") +
	scale_y_continuous("Uni degree") +
	theme_bw())

```



```{r, fig.height = 6}
tm_shape(NLD_muni) +
  tm_polygons("edu_appl_sci",
  			fill.legend = tm_legend("University degree")) +
tm_inset(g1, position = c("left", "top"))
```

In the next example we add a scatter plot and another tmap plot, also added via `tm_inset()`.

Both insets plots will be minimal, so without legend: 

```{r, fig.height = 3}
(g2 = ggplot(NLD_muni, aes(x = income_high, y = edu_appl_sci, color = urbanity)) +
	geom_point(size = 2) +
 	scale_color_brewer(guide = "none", type = "div") +
	theme_bw())
```

```{r, fig.height = 3}
(tm = tm_shape(NLD_muni) +
	tm_dots(fill = "urbanity",
			fill.scale = tm_scale(values = "brewer.br_bg"),
			fill.legend = tm_legend_hide()) +
	tm_layout(frame = FALSE))
```

Putting it all together

```{r, fig.height = 6}
tm_shape(NLD_muni) +
  tm_polygons("edu_appl_sci",
  			fill.legend = tm_legend("University degree", group_id = "right")) +
tm_inset(g1, group_id = "right", height = 10, width = 10) +
tm_inset(g2, group_id = "right", height = 10, width = 10) +
tm_inset(tm, group_id = "topleft", position = c("left", "top"), box_frame = FALSE) +
tm_components(group_id = "right", position = tm_pos_out("right", "center"))
```
