---
title: "tmap advanced: grouping of components"
output: 
  bookdown::html_vignette2:
pkgdown:
  as_is: true
template:
  math-rendering: mathjax
bibliography: '`r system.file("tmap.bib", package="tmap")`'
csl: "`r system.file('ieee.csl', package = 'tmap')`"
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  out.width = "100%",
  dpi = 300,
  fig.width = 7.2916667,
  comment = "#>"
)
hook_output <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

```


```{r, echo = FALSE, message = FALSE}
library(tmap)
tmap_options(scale = 0.75)
```

### Note: this is written for the upcoming version 4.1

When component share the same position (see [vignette about positions](https://r-tmap.github.io/tmap/articles/adv_positions)), we can specify how they are grouped via `tm_comp_group()`.

We will use a self-made map component, made with the grid package that we show via `tm_inset()`.
```{r, fig.height = 6}
g = function(text) grid::gList(
  grid::rectGrob(gp=grid::gpar(fill = "gold")), 
  grid::circleGrob(r = .45), 
  grid::textGrob(text))
b = function(text) grid::gList(
  grid::rectGrob(gp=grid::gpar(fill = "steelblue")), 
  grid::circleGrob(r = .45), 
  grid::textGrob(text))
```

# Linking components via `group_id`

```{r, fig.height = 6}
tm = tm_shape(NLD_muni) +
	tm_polygons("grey95", col = "grey90") +
tm_inset(g("A1"), group_id = "A") +
tm_inset(g("A2"), group_id = "A") +
tm_inset(b("B1"), group_id = "B", height = 3, width = 3) +
tm_inset(g("A3"), group_id = "A") +
tm_inset(b("B2"), group_id = "B", height = 3, width = 3) +
tm_layout(asp = 1)

tm + 
  tm_comp_group("A", position = tm_pos_in("left", "top")) +
  tm_comp_group("B", position = tm_pos_in("right", "bottom"))
```

Note that the actual grouping is done by position alone: all components with the same position are considered as a group and therefore stacked instead of drawn on top of each other.
The goal of `tm_comp_group()` is only to specify position and group related arguments (which we discuss in this vignette) directly for multiple components.

In this example components of A and B have the same position and are therefore grouped:

```{r, fig.height = 6}
tm + 
  tm_comp_group("A", position = tm_pos_in("right", "bottom")) +
  tm_comp_group("B", position = tm_pos_in("right", "bottom"))
```


# Stacking

## Direction

The `stack` argument is either `"horizontal"` or `"vertical"`:

```{r, fig.height = 6}
tm + 
  tm_comp_group("A", position = tm_pos_in("left", "top"), stack = "horizontal") +
  tm_comp_group("B", position = tm_pos_in("right", "bottom"), stack = "vertical")
```

## Order

The stacking order is by default the same as in which the components appear, but can be changed with the `z` argument in each component function:

```{r, fig.height = 6}
tm_shape(NLD_muni) +
  tm_polygons("grey95", col = "grey90") +
tm_inset(g("A1"), group_id = "A", z = 1) +
tm_inset(g("A2"), group_id = "A", z = 3) +
tm_inset(g("A3"), group_id = "A", z = 2) +
tm_layout(asp = 1) +
tm_comp_group("A", position = tm_pos_in("left", "top")) 
```

# Frame

## `frame_combine`

By default `frame_combine = TRUE`, which means that a single frame is drawn for all components in the same group. This can be disabled with `frame_combine = FALSE`


```{r, fig.height = 6}
tm_shape(NLD_muni) +
  tm_polygons("grey95", col = "grey90") +
tm_inset(g("A1"), group_id = "A") +
tm_inset(g("A2"), group_id = "A", width = 3, height = 3) +
tm_inset(g("A3"), group_id = "A") +
tm_layout(asp = 1) +
tm_comp_group("A", position = tm_pos_in("left", "top"), frame_combine = FALSE) 
```

## `equalize`

Equalize (by default `TRUE`) means that components made equal in size, that is: the widths in case they are stacked vertically (as in this example), and the heights in case they are stacked horizontally.

```{r, fig.height = 6}
tm_shape(NLD_muni) +
  tm_polygons("grey95", col = "grey90") +
tm_inset(g("A1"), group_id = "A") +
tm_inset(g("A2"), group_id = "A", width = 3, height = 3) +
tm_inset(g("A3"), group_id = "A") +
tm_layout(asp = 1) +
tm_comp_group("A", position = tm_pos_in("left", "top"), frame_combine = FALSE, equalize = FALSE) 
```

# Margins

## General `offset` and `stack_margin`

The `offset` specifies the distance (in line heights) between the frame and components. By default 0.75 in case the components are drawn inside the frame and `pos.h` and `pos.v` are lower case, and 0 otherwise.

The `stack_margin` specifies the distance between components. In case `frame_combine` it is 0.5 by default, otherwise 0.

```{r, fig.height = 6}
tm_shape(NLD_muni) +
  tm_polygons("grey95", col = "grey90") +
tm_inset(g("A1"), group_id = "A") +
tm_inset(g("A2"), group_id = "A", width = 3, height = 3) +
tm_inset(g("A3"), group_id = "A") +
tm_layout(asp = 1) +
tm_comp_group("A", position = tm_pos_in("left", "top"), frame_combine = FALSE, equalize = FALSE, offset = 0, stack_margin = 1)
```

Note: currently both `offset` and `stack_margin` are applied for all components for which `pos.h` and `pos.v` are specified as characters.


## Component specific `margins`

Each component has its own margins, which are the margins to the frame. These can be specified with a vector of four: bottom, left, top, right.

```{r, fig.height = 6}
tm_shape(NLD_muni) +
  tm_polygons("grey95", col = "grey90") +
tm_inset(g("A1"), group_id = "A", margins = 0) +
tm_inset(g("A2"), group_id = "A", width = 3, height = 3, margins = 0) +
tm_inset(g("A3"), group_id = "A", margins = 0) +
tm_layout(asp = 1) +
tm_comp_group("A", position = tm_pos_in("left", "top"), frame_combine = T, frame = TRUE, offset = 0, stack_margin = 0)
```

# Alignment

Recall that the `align.h` and `align.v` arguments of `tm_pos()` specify the alignment of the components [see vignette about positions](https://r-tmap.github.io/tmap/articles/adv_positions)).
This can be set for each component individually, or, in case it is the same for all components inside `tm_comp_group()`.

```{r, fig.height = 6}
tm_shape(NLD_muni) +
  tm_polygons("grey95", col = "grey90") +
tm_inset(g("A1"), group_id = "A") +
tm_inset(g("A2"), group_id = "A", width = 3, height = 3) +
tm_inset(g("A3"), group_id = "A") +
tm_layout(asp = 1) +
tm_comp_group("A", position = tm_pos_in("left", "top", align.h = "right"), frame_combine = FALSE, equalize = FALSE, offset = 0, stack_margin = 0) 
```

# Frame and background color and alpha

```{r, fig.height = 6}
tm_shape(NLD_muni) +
  tm_polygons("grey75", col = "grey90") +
tm_inset(g("A1"), group_id = "A") +
tm_inset(g("A2"), group_id = "A", width = 3, height = 3) +
tm_inset(g("A3"), group_id = "A") +
tm_layout(asp = 1) +
tm_comp_group("A", position = tm_pos_in("left", "bottom"), frame = TRUE, bg.color = "purple", bg.alpha = 0.5)
```
