---
title: "tmap advanced: positioning of components"
output: 
  bookdown::html_vignette2:
pkgdown:
  as_is: true
template:
  math-rendering: mathjax
bibliography: '`r system.file("tmap.bib", package="tmap")`'
csl: "`r system.file('ieee.csl', package = 'tmap')`"
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  out.width = "100%",
  dpi = 300,
  fig.width = 7.2916667,
  comment = "#>"
)
hook_output <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

```

```{r, echo = FALSE, message = FALSE}
library(tmap)
tmap_options(scale = 0.75)
```

Recall from the [vignette about components](https://r-tmap.github.io/tmap/articles/basics_components) that the `position` argument of map components specifies their position.

It can be specified in two ways:

* An object from the `tm_pos()`, or one of its wrappers, such as `tm_pos_in()` and `tm_pos_out()`.
* A vector of length two, which is a shortcut for placement inside the map frame, a shortcut for `tm_pos_in()`

We will use a self-made map component, made with the grid package that we show via `tm_inset()`.
```{r, fig.height = 6}
g = function(text) grid::gList(
  grid::rectGrob(gp=grid::gpar(fill = "gold")), 
  grid::circleGrob(r = .45), 
  grid::textGrob(text))
b = function(text) grid::gList(
  grid::rectGrob(gp=grid::gpar(fill = "steelblue")), 
  grid::circleGrob(r = .45), 
  grid::textGrob(text))
```

# Shortcut method (inside the map frame)

The `position` can be a vector of two values, where the first is the horizontal position and the second the vertical position. It is a shortcut for `tm_pos_in()`, which is explained in the next sections.

There are two options:

## Character vector of two

* the first value specifies the horizontal position (`"left"`, `"center"`, `"right"`), and 
* the second one the vertical position (`"top"`, `"center"`, `"bottom"`). 

Upper case means tight to the frame. Otherwise an offset margin is applied (can be specified with `offset`, see below).

## Numeric vector of two
	
This can be seen as the 'manual' mode, because it provides more freedom. 

* the first value specifies the horizontal position, where 0 corresponds to left and 1 to right.
* the second value specifies the vertical position where 0 is bottom and 1 top. 

By default, these values will specify the position of the **left top corner** of the component. This justification point can be changed using the `just.h` and `just.v` arguments of [tm_pos()], see below.

## Example

The character positioned components are in gold and the numeric positioned components in blue.

```{r, fig.height = 6}
tm_shape(NLD_muni) +
	tm_polygons("grey85", col = "grey60") +
	tm_inset(g("left, top"), position = c("left", "top")) +
	tm_inset(g("right, bottom"), position = c("right", "bottom")) +
	tm_inset(g("LEFT, BOTTOM"), position = c("LEFT", "BOTTOM")) +
	tm_inset(b("0.3, 0.3"), position = c(0.2, 0.2))	+
	tm_inset(b("0.5, 0.5"), position = c(0.5, 0.5)) +		
	tm_inset(b("0.9, 1"), position = c(0.9, 1))		
```

# Advanced method with `tm_pos()`

## The 3 x 3 grid

The whole plot area is a 3 x 3 grid, where the map (or facets) are drawn in the center grid cell.

There are three variants (wrappers) of the main function `tm_pos()`:

* `tm_pos_out()` draws the component outside the map frame (so in one of the 3 x 3 grid cells)
* `tm_pos_in()` and `tm_pos_on_top()` draw the component in the middle grid cell. The difference is that the former takes the frame (and offset from it) into account while the latter doesn't.


## `cell.h` and `cell.v`

The arguments `cell.h` and `cell.v` determine the grid cell. These should only be used for `tm_pos_out()`. For the other position function, they are already set to `"center"` (the center grid cell in which the map is drawn).


```{r, fig.height = 6}
tm_shape(NLD_muni) +
	tm_polygons("grey85", col = "grey60") +
tm_inset(g("center, center"), position = tm_pos_on_top(pos.h = "center", pos.v = "center")) +
tm_inset(g("left, center"), position = tm_pos_out("left", "center", pos.v = "center")) +
tm_inset(g("right, center"), position = tm_pos_out("right", "center", pos.v = "center")) +
tm_inset(g("left, top"), position = tm_pos_out("left", "top")) +
tm_inset(g("center, top"), position = tm_pos_out("center", "top", pos.h = "center")) +
tm_inset(g("right, top"), position = tm_pos_out("right", "top")) +
tm_inset(g("left, bottom"), position = tm_pos_out("left", "bottom")) +
tm_inset(g("center, bottom"), position = tm_pos_out("center", "bottom", pos.h = "center")) +
tm_inset(g("right, bottom"), position = tm_pos_out("right", "bottom")) +
tm_layout(asp = 1, frame.color = "blue", frame.lwd = 2)
```

Note that we used `pos.h` and `pos.v`. These will be explained in the next section.

## `pos.h` and `pos.v`

The position of a component within a cell. These values are similar to the shortcut values described above:

* Character values: `"left"`, `"center"`, or `"right"` (or in upper case) for `pos.h` and `"top"`, `"center"`, and `"bottom"` for `pos.v`.
* Numeric values: 0 (left) to 1 (right) for `pos.h`, and 0 (bottom) to 1 (top) for `pos.v`

```{r, fig.height = 6}
tm_shape(NLD_muni) +
  tm_polygons("grey85", col = "grey60") +
tm_inset(g("pos.v = \"top\" in\ncell (left, center)"), 
  position = tm_pos_out("left", "center", pos.v = "top")) +
tm_inset(g("pos.v = \"bottom\" in\ncell (right, center)"), 
  position = tm_pos_out("right", "center", pos.v = "bottom")) +
tm_inset(b("pos.h = 0.0\npos.v = 0.0\nin center cell"),
  position = tm_pos_in(pos.h = 0.0, pos.v = 0.0)) +
tm_inset(b("pos.h = 0.5\npos.v = 0.5\nin center cell"),
  position = tm_pos_in(pos.h = 0.5, pos.v = 0.5)) +
tm_inset(b("pos.h = 1\npos.v = 1\nin center cell"),
  position = tm_pos_in(pos.h = 1, pos.v = 1)) +
tm_layout(asp = 1, frame.color = "blue", frame.lwd = 2)
```

Recall that the **justification point** for the numeric specifications is the **top left corner** of the component. This can be adjusted via `just.h` and `just.v` as explained next.

## `just.h` and `just.v`

In case `pos.h` and `pos.v` are numbers, the justification can be set with

* `just.h`: `"left"`, `"center"`, or `"right"`
* `just.v`: `"top"`, `"center"`, and `"bottom"`

```{r, fig.height = 6}
tm_shape(NLD_muni) +
  tm_polygons("grey85", col = "grey60") +
tm_inset(b("center, center"), 
  position = tm_pos_in(0.5, 0.5, just.h = "center", just.v = "center")) +
tm_inset(b("left, top"), 
  position = tm_pos_in(0.5, 0.5, just.h = "left", just.v = "top")) +
tm_inset(b("right, bottom"), 
  position = tm_pos_in(0.5, 0.5, just.h = "right", just.v = "bottom")) +
tm_layout(asp = 1, frame.color = "blue", frame.lwd = 2)
```



## `align.h` and `align.v`

In case there are multiple components in the same position (see next section), the alignment determines how a component is aligned.


The alignment of components within the same cell:

```{r, fig.height = 4}
tm_shape(NLD_muni) +
  tm_polygons("grey85", col = "grey60") +
tm_inset(g(""), 
  width = 4, height = 4,
  position = tm_pos_in("left", "top")) +
tm_inset(b("left"),
  width = 3, height = 3,
  position = tm_pos_in("left", "top", align.h = "left")) +
tm_inset(b("right"),
  width = 3, height = 3,
  position = tm_pos_in("left", "top", align.h = "right")) +
tm_inset(g(""), 
  width = 4, height = 4,
  position = tm_pos_in("left", "top")) +
tm_layout(asp = 1, frame.color = "blue", frame.lwd = 2)
```

## Automatic positioning

The default position of legends and map components are in some cases done automatically. Automatic positioning can be achieved by setting a `position` argument to `tm_pos_auto_in()` or `tm_pos_auto_out()`. The former checks in which map corner is most space available (and bases `pos.h` and `pos.v` on that). The latter calculates in which grid cell (so `cell.h` and `cell.v`) the map components can be positioned, which is based on the aspect ratio and margins.

# Grouping components

What happens when components share the same position? See [next vignette](https://r-tmap.github.io/tmap/articles/adv_comp_group)
