---
title: "tmap advanced: margins and aspect ratio"
output: 
  bookdown::html_vignette2:
pkgdown:
  as_is: true
template:
  math-rendering: mathjax
bibliography: '`r system.file("tmap.bib", package="tmap")`'
csl: "`r system.file('ieee.csl', package = 'tmap')`"
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  out.width = "100%",
  dpi = 300,
  fig.width = 7.2916667,
  comment = "#>"
)
hook_output <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

```


```{r, echo = FALSE, message = FALSE}
library(tmap)
tmap_options(scale = 0.75)
```


## Design mode

tmap offers a special **design mode**, which can be enabled via `tmap_design_mode`. Bolow the same map, but the second time with this design mode enabled:

```{r, figures-side, fig.show="hold", out.width="47%", fig.height=8}
(tm = tm_shape(NLD_muni) +
  tm_polygons(fill = "employment_rate", 
    fill.legend = tm_legend(position = tm_pos_in("left", "top"))) +
   tm_compass(position = tm_pos_in("right", "bottom")) +
   tm_scalebar(position = tm_pos_in("left", "bottom")))

tmap_design_mode()

tm
```


## tm_layout() and tmap options

All options related to the layout of the map, such as background color and margins, can be set via `tm_layout`.

These options are a subset of all tmap options, which can be set via `tmap_options` (a stand-alone function) or `tm_options`, which is supposed to be stacked with the `+` operator.

## Inner and outer margins

Inner and outer margins can be set with the options `inner.margins` and `outer.margins`. 

The inner margins are the margins between the shape (i.e. spatial geometries) and the map frame. In the design mode, the shape area with the inner margins is shown with a yellow rectangle.




```{r, fig.height=5}
tm_shape(NLD_muni) +
  tm_polygons(fill = "employment_rate") +
	tm_layout(
		inner.margins = c(0.2, 0.2, 0.2, 0.2))
```

This vector of four numbers correspond to respectively the bottom, left, top, and right margin.

The outer margins are the margins between the map frame and the device frame.


```{r, fig.height=5}
tm_shape(NLD_muni) +
  tm_polygons(fill = "employment_rate") +
	tm_layout(
		outer.margins = c(0.2, 0.2, 0.2, 0.2))
```

## Meta margins

Besides inner and outer margins, there is another type of margins: the meta margins. This space dedicated for map components that are outside the map (see [vignette about component   positions](https://r-tmap.github.io/tmap/articles/adv_positions#cell-h-and-cell-v). For single maps this is calculated automatically, using the other margins and the map component sizes. If there are no outside map components, `meta.margins` is set to `c(0, 0, 0, 0)`. For complex maps (e.g. facetted maps) it is set to 0.4 for each side in which map components are placed.

## Aspect ratio

The aspect of the map frame can be set via the option `asp`. If so, the inner margins will be adjusted automatically.

```{r, fig.height=5}
tm_shape(NLD_muni) +
  tm_polygons(fill = "employment_rate") +
	tm_title("Square map") +
	tm_layout(asp = 1)
```

The special value 0 means that the aspect ratio is adjusted to meet the space that is left on the device after substracting the outer margins and space occupied by the outside components (in this case just the legends).

```{r, fig.height=5}
tm_shape(NLD_muni) +
  tm_polygons(fill = "employment_rate") +
	tm_layout(outer.margins = c(0.2, 0.2, 0.2, 0.2),
			  asp = 0)
```

