---
title: "tmap advanced: inset maps and minimaps"
output: 
  bookdown::html_vignette2:
bibliography: '`r system.file("tmap.bib", package="tmap")`'
csl: "`r system.file('ieee.csl', package = 'tmap')`"
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 8,
  comment = "#>"
)
hook_output <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

```


```{r, echo = FALSE, message = FALSE}
library(tmap)
tmap_options(scale = 0.75)
```

## Introduction

An inset map or minimap is a small map embedded in the main map. The purpose of such a small map is 1) to show geographic context (zoomed out) or 2) to show details (zoomed in). The first type is called a minimap, created via `tm_minimap`. The second type can be created with the general purpose function `tm_inset()`. So

* we use `tm_minimap(()` to show a broader geographic context; where on earth is the main map located?
* we use `tm_inset()` to zoom in on areas that are too cluttered at the main maps's scale.

Note that both `tm_inset()` and `tm_minimap()` are map components (see [vignette](https://r-tmap.github.io/tmap/articles/basics_components))

## Minimaps


```{r, fig.height = 6}
tmap_mode("plot")
tm_shape(NLD_muni) +
	tm_polygons(
		fill = "edu_appl_sci") +
	tm_title("Population share with (applied) university degree in 2022") +
	tm_credits("Statistics Netherlands (CBS)") + 
	tm_minimap(position = c("left", "top"))
```

The implementation in plot mode is experimental (and new as of tmap 4.2). It features a small globe. There are no options implemented yet to change the appearance.
Note that it takes a few seconds to run due to the applied orthographic [map projection](https://r-tmap.github.io/tmap/articles/foundations_crs).

The minimap is interactive in view mode.

```{r, fig.height = 6}
tmap_mode("view")
tmap_last()
```

Note that the minimap is also supported in two other modes, `"maplibre"` and `"mapbox"` via the extension package [**tmap.mapgl**](https://r-tmap.github.io/tmap.mapgl/).


## Inset maps

Inset (zoomed in) maps are only implemented in `"plot"` mode, because this feature is less usefull in interactive modes (`"view"`, `"maplibre"` and `"mapbox"`) because of the user's ability to zoom in.

The function `tm_inset()` is general purpose. It can accept many input objects including ggplot2 objects and image files. In this vignette, we look at two types of inputs: a bounding box and a tmap object.

### The same map zoomed in

When a bounding box is provided, the same map is plotted, but using the provided bounding box. This bounding box is usually smaller than the bounding box of the main map in order to zoom in on a region of interest.

The bounding box of the inset map can be a `bbox` object created with `sf::st_bbox()` or `tmaptools::bb()`. The latter can also be used to query Open Street Map; if a character is provided, it will use `tmaptools::geocode_OSM` to look for the corresponding bounding box.

```{r, fig.height = 6}
library(tmaptools)

tmap_mode("plot")
tm_shape(NLD_dist) +
	tm_polygons(
		fill = "edu_appl_sci",
		col = NULL) +
tm_shape(NLD_muni) +
	tm_borders() +
tm_title("Population share with (applied) university degree in 2022") +
tm_credits("Statistics Netherlands (CBS)") + 
tm_title("Amsterdam", size = 0.8, position = c("left", "top")) +
tm_inset(bb("Amsterdam"), position = c("left", "top")) +
tm_title("The Hague", size = 0.8, position = c("left", "top")) +
tm_inset(bb("The Hague"), position = c("left", "top")) +
tm_layout(inner.margins = c(0.05, 0.2, 0.05, 0.05))
```



## Custom inset maps

To have more flexibility it is also possible to inset a different tmap object.

```{r, fig.height = 6}
NLD_muni$pop_est_dens = NLD_muni$population / NLD_muni$area

tm_NLD = tm_shape(NLD_muni) +
	tm_polygons(fill = "pop_est_dens", 
				fill.legend = tm_legend_hide(),
				fill.scale = tm_scale_continuous_sqrt(values = "viridis", limits = c(0, 6500)),
				col = NULL) +
	tm_shape(NLD_prov) +
	tm_borders()

tm_shape(World) +
	tm_polygons(fill = "pop_est_dens", 
				col = "black",
				fill.scale = tm_scale_continuous_sqrt(values = "viridis", limits = c(0, 6500))) +
	tm_inset(tm_NLD, position = c("left", "bottom")) +
	tm_crs("+proj=robin")
```
