---
title: "tmap basics: charts"
output: 
  bookdown::html_vignette2:
pkgdown:
  as_is: true
template:
  math-rendering: mathjax
bibliography: '`r system.file("tmap.bib", package="tmap")`'
csl: "`r system.file('ieee.csl', package = 'tmap')`"
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  out.width = "100%",
  dpi = 300,
  fig.width = 7.2916667,
  comment = "#>"
)
hook_output <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

```


```{r, echo = FALSE, message = FALSE}
library(tmap)
tmap_options(scale = 0.8)
```



## Introduction

Each visual variable (e.g. `fill` in `tm_polygons()`) has an additional `.chart` argument via which charts can be shown:

```{r, fig.height = 5}
tm_shape(World) +
  tm_polygons(
    fill = "press",
    fill.scale = tm_scale_intervals(n=10, values = "scico.hawaii"),
    fill.legend = tm_legend("World Press\nFreedom Index"),
	fill.chart = tm_chart_bar()) +
tm_crs("auto")
```

## Chart types

### Numeric variables 

```{r, fig.height = 5}
tm_shape(World) +
  tm_polygons("HPI",
    fill.scale = tm_scale_intervals(),
    fill.chart = tm_chart_donut())
```


```{r, fig.height = 5}
tm_shape(World) +
  tm_polygons("HPI",
    fill.scale = tm_scale_intervals(),
    fill.chart = tm_chart_box())
```

```{r, fig.height = 5}
tm_shape(World) +
  tm_polygons("HPI",
    fill.scale = tm_scale_intervals(),
    fill.chart = tm_chart_violin())
```

### Categorical variable

```{r, fig.height = 5}
tm_shape(World) +
  tm_polygons("economy",
    fill.scale = tm_scale_categorical(),
    fill.chart = tm_chart_bar())
```

```{r, fig.height = 5}
tm_shape(World) +
  tm_polygons("economy",
    fill.scale = tm_scale_categorical(),
    fill.chart = tm_chart_donut())
```

### Bivariate charts

```{r, fig.height = 5}
tm_shape(World) +
  tm_polygons(tm_vars(c("HPI", "well_being"), multivariate = TRUE),
    fill.chart = tm_chart_heatmap())
```

## Position

We can update the position of the chart to bottom right (in a separate frame). See [vignette about positioning](https://r-tmap.github.io/tmap/articles/adv_positions).

```{r, fig.height = 6}
tm_shape(World) +
  tm_polygons(
    fill = "press",
    fill.scale = tm_scale_intervals(n=10, values = "scico.hawaii"),
    fill.legend = tm_legend("World Press\nFreedom Index"),
	fill.chart = tm_chart_bar(position = tm_pos_out("center", "bottom", pos.h = "right"))) +
tm_crs("auto")
```

Or, in case we would like the chart to be next to the legend, but in a different frame:

```{r, fig.height = 6}
tm_shape(World) +
  tm_polygons(
    fill = "press",
    fill.scale = tm_scale_intervals(n=10, values = "scico.hawaii"),
    fill.legend = tm_legend("World Press\nFreedom Index", group.frame = FALSE),
	fill.chart = tm_chart_bar(position = tm_pos_out("center", "bottom", align.v = "top"))) +
	tm_layout(component.stack_margin = .5) +
tm_crs("auto")
```

## Additional ggplot2 code

```{r, fig.height = 6}
require(ggplot2)
tm_shape(World) +
  tm_polygons("HPI",
    fill.scale = tm_scale_intervals(),
    fill.chart = tm_chart_bar(
      extra.ggplot2 = theme(
        panel.grid.major.y = element_line(colour = "red")
      ))
    )
```
