---
title: "tmap basics: modes"
output: 
  bookdown::html_vignette2:
bibliography: '`r system.file("tmap.bib", package="tmap")`'
csl: "`r system.file('ieee.csl', package = 'tmap')`"
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 8,
  comment = "#>"
)
hook_output <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

```


```{r, echo = FALSE, message = FALSE}
library(tmap)
tmap_options(scale = 0.75)
```



## Modes

tmap facilitates two output modes: `"plot"`, which produces static maps, and `"view"` which produces (using the same tmap code) interactive maps. As of version 4, tmap can also be extended with other modes, as demonstrated below.

 
## Switching between modes

```{r}
## current mode
tmap_mode()

## to view mode
tmap_mode("view")

## back to plot mode
tmap_mode("plot")
```

The handy function `ttm()` is used to toggle between the modes.


## Plot mode

We start with cerating the plot and assign it to a variable called `tm`.

```{r}
tm = tm_shape(World, crs = 8857) +
  tm_polygons(
    fill = "press",
    fill.scale = tm_scale_intervals(values = "pu_gn")) +
tm_shape(metro) +
  tm_bubbles(
    size = "pop2020",
    fill = "gold",
    size.scale = tm_scale_continuous(values.scale = 0.8, n = 8))
```

We are in `"plot"` mode. Now we can plot the map by printing `tm` object:

```{r, fig.height = 3.5, out.width = "100%", dpi = 300, fig.width = 7.2916667}
tm
```


## View mode

The same map in view mode:

```{r}
ttm()
tm
```

Note that there is big difference: in this `"view"` mode there are basemaps, and in `"plot"` mode none. This is caused by different default option.

Basemaps can be enabled or disabled via `tm_basemap()`: 

```{r, fig.height = 3.5}
tm + tm_basemap(NULL)
```

See [vignette about basemaps](https://r-tmap.github.io/tmap/articles/06_basics_basemaps).

## Mode specific options

Mode specific layout options can be set via `tm_plot()` and `tm_view()`. The number of options in `tm_plot()` is limited to just one, because it uses all general purpose options. In contrast, `tm_view()` contains more options, e.g. there to position the control box and what the default zoom level is:

```{r, fig.height = 3.5}
tm +
	tm_view(control.position = c("left", "bottom"),
			set_view = 2)
```

For a more detailed description of options, see [vignette about options](https://r-tmap.github.io/tmap/articles/42_adv_options).


## Other modes

As of version 4, tmap can be extended with other modes. See ([vignette about extensions](https://r-tmap.github.io/tmap/articles/44_adv_extensions)).


```{r, message=FALSE}
library(tmap.deckgl)
tmap_mode("deck")
tm
```

The proof-of-concept package `tmap.deckgl` only features the map layer function `tm_polygons()` (so the bubbles are not working yet).


