---
title: "tmap basics: legends"
output: 
  bookdown::html_vignette2:
pkgdown:
  as_is: true
template:
  math-rendering: mathjax
bibliography: '`r system.file("tmap.bib", package="tmap")`'
csl: "`r system.file('ieee.csl', package = 'tmap')`"
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  out.width = "100%",
  dpi = 300,
  fig.width = 7.2916667,
  comment = "#>"
)
hook_output <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

```


```{r, echo = FALSE, message = FALSE}
library(tmap)
tmap_options(scale = 0.75)
```



### About the data

A spatial data object contained in tmap is called `World`. It is a data frame with a row for each country. The columns are the following data variables plus an additional geometry column which contains the geometries (see sf package):

```{r}
names(World)
```

We specify this object with `tm_shape` (see other vignette) and for convenience assign it to `s`:

```{r}
s = tm_shape(World, crs = "+proj=eqearth")
```


## Legend type

Each visual variable, e.g. `fill` in `tm_polygons` will by default produce a legend. The type of legend depends on used scale (see \href{https://r-tmap.github.io/tmap/articles/basics_scales}{vignette about scales}). E.g. `tm_scale_continuous` applied to `fill` will produce a continuous color gradient:

```{r, fig.height = 3.5}
s + tm_polygons(
  fill = "HPI",
  fill.scale = tm_scale_continuous(values = "pu_gn"))
```

For `tm_scale_intervals`, there are two styles. A choice can be made via the argument `label.style`. By default it is set to discrete:

```{r, fig.height = 3.5}
s + tm_polygons(
  fill = "HPI",
  fill.scale = tm_scale_intervals(values = "pu_gn"))
```

Alternatively, it can be set to `"continuous"`:

```{r, fig.height = 3.5}
s + tm_polygons(
  fill = "HPI",
  fill.scale = tm_scale_intervals(values = "pu_gn", label.style = "continuous"))
```

## `.legend` arguments

Each visual variable also has a `.legend` argment, such as `fill.legend` in `tm_polygons`. This can be used to:

* set the title of the legend,
* change the orientation of the legend (portrait or landscape),
* reverse the legend items,
* hide the legend, and
* change the size and position of the legend.

Note that the content of the legend, so breaks, scales, and ticks are related to the scale and should therefore be specified in the scale functions.

## Title, orientation and size

The following example illustrates how the title, orientation and size and specified: 

```{r, fig.height = 3.5}
s + tm_polygons(
  fill = "HPI",
  fill.scale = tm_scale_continuous(values = "pu_gn"),
  fill.legend = 
    tm_legend(
      title = "Happy Planex Index", 
      orientation = "landscape", 
      width = 60))
```

The legend width is set in number of text line heights, which corresponds to the number of characters. In other words, the resulting width depends on the font size, which can be set in `tm_legend`, but also via the general `scale` option, which determines the overall 'scale' of the map (all line widths, point sizes and font sizes):

```{r, fig.height = 3.5}
s + tm_polygons(
  fill = "HPI",
  fill.scale = tm_scale_continuous(values = "pu_gn"),
  fill.legend = 
    tm_legend(
      title = "Happy Planex Index", 
      orientation = "landscape", 
      width = 60)) +
tm_layout(scale = 0.25)
```

## Hiding legends

A legend can be suppressed by setting the `.legend` argument to `tm_legend_hide()`.

```{r, fig.height = 3.5}
s + tm_polygons(
  fill = "HPI",
  fill.scale = tm_scale_continuous(values = "pu_gn"),
  fill.legend = tm_legend_hide())
```

See the [advanced vignette](https://r-tmap.github.io/tmap/articles/adv_legends) for more options.
