---
title: "tmap basics: layout"
output: 
  bookdown::html_vignette2:
pkgdown:
  as_is: true
template:
  math-rendering: mathjax
bibliography: '`r system.file("tmap.bib", package="tmap")`'
csl: "`r system.file('ieee.csl', package = 'tmap')`"
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  out.width = "100%",
  dpi = 300,
  fig.width = 7.2916667,
  comment = "#>"
)
hook_output <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

```


```{r, echo = FALSE, message = FALSE}
library(tmap)
tmap_options(scale = 0.75)
```

## What is layout?

With layout we mean all aspects of how the plot looks like, except for 

* specifications of data-driven [visual variables](https://r-tmap.github.io/tmap/articles/basics_vv).
* the layout of [legends](https://r-tmap.github.io/tmap/articles/basics_legends#layout)
* the layout of [map components](https://r-tmap.github.io/tmap/articles/basics_components)

What is left? Background colors, frames, panels, fonts, margins, etc. All these layout options can be changed via `tm_layout()`. 

In this vignette, we'll cover the most important layout settings. First, let's create a map:

```{r}
tm = tm_shape(World, crs = "+proj=eqearth") + tm_polygons("HPI", fill.scale = tm_scale_intervals(values = "pu_gn"))
```


## Background colors

The background colors inside and outside the map frame are specified as follows:

```{r, fig.height=3.5}
tm + tm_layout(
  bg.color = "skyblue",
  outer.bg.color = "gold")
```

The backgrounds can be disabled by setting `bg` and `outer.bg` (inside and outside the frame respectively) to `FALSE`. Note that this does not change the layout of the map; this is accomplished via the margins (see below).



## Frame

The map frame can be disabled by setting `frame = FALSE`:

```{r, fig.height=3.5}
tm + tm_layout(
  bg.color = "grey90",
  frame = FALSE)
```


## Earth boundaries

For certain [map projections](https://r-tmap.github.io/tmap/articles/foundations_crs), including the used one, we can infer the 'earth boundaries'. We can enable them with the option `earth_boundary`. The background color outside of the earth boundaries (and inside the map frame if specified) are determined by `space.color`.

For this type of map, it makes sense to disable the map frame, and place the legend in the corner of the map, perhaps even with a bit of overlap.

```{r, fig.height=3.5}
tm_shape(World, crs = "+proj=eqearth") + 
  tm_polygons(
    fill = "HPI", 
    fill.scale = tm_scale_intervals(values = "pu_gn"),
    fill.legend = tm_legend(position = c("left", "bottom"))) +
tm_layout(bg.color = "skyblue",
  earth_boundary = TRUE,
  frame = FALSE,
  space.color = "white")
```

## Panels

To change the appearance of panels the options with the prefix `panel.` are used:

```{r, fig.height = 7}
tm_shape(World, crs = "+proj=eqearth") +
  tm_polygons(c("well_being", "footprint")) +
tm_layout(panel.label.bg.color = "gold",
		  panel.label.size = 2,
		  panel.label.height = 3)	
```

Panels can be disabled using `panel.show = FALSE`:

```{r, fig.height = 7}
tm_shape(World, crs = "+proj=eqearth") +
  tm_polygons(c("well_being", "footprint")) +
tm_layout(panel.show = FALSE)	
```

## Margins

Margins can be set with `inner.margins`, `outer.margins`, `meta.margins`. The inner margins are the margins between map and the frame

```{r, fig.height = 3.5}
tm +
  tm_layout(inner.margins = c(0, 0, 0.02, 0.02))	
```

The four numbers are the margins for bottom, left, top, and right respectively. The units are relative to the map frame, so 0.02 means (about) 2 percent of the frame height.

The outer margins are the margins between map frame and plotting device:

```{r, fig.height = 3.5}
tm_shape(World, crs = "+proj=eqearth") + 
	tm_polygons("HPI", 
				fill.scale = tm_scale_intervals(values = "pu_gn"),
				fill.legend = tm_legend(position = c("left", "bottom"))) +
  tm_layout(outer.margins = c(0, 0, 0, 0), asp = 0)
```

In this example we also set `asp = 0` meaning that the map frame should fit inside the graphics device. That means that the inner margins are increased if the aspect ratio (width/height) of the graphics device is different than that of the map. Setting the margins can be quite difficult and is explained in detail in this [ vignette](https://r-tmap.github.io/tmap/articles/adv_margins).
