---
title: "tmap basics: scales"
output: 
  bookdown::html_vignette2:
pkgdown:
  as_is: true
template:
  math-rendering: mathjax
bibliography: '`r system.file("tmap.bib", package="tmap")`'
csl: "`r system.file('ieee.csl', package = 'tmap')`"
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  out.width = "100%",
  dpi = 300,
  fig.width = 7.2916667,
  comment = "#>"
)
hook_output <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

```


```{r, echo = FALSE, message = FALSE}
library(tmap)
tmap_options(scale = 0.75)
```



### About the data

A spatial data object contained in tmap is called `World`. It is a data frame with a row for each country. The columns are the following data variables plus an additional geometry column which contains the geometries (see sf package):

```{r}
names(World)
```

We specify this object with `tm_shape` (see other vignette) and for convenience assign it to `s`:

```{r}
s = tm_shape(World, crs = "+proj=eqearth")
```


## Scales: numeric data (intervals)

Each visual variable, e.g. `fill` in `tm_polygons` can represent a data variable:

```{r, fig.height = 3.5}
s + tm_polygons(fill = "HPI")
```

A *scale* defines how to map the data values to visual values. Numeric data variables (e.g. `"HPI"` which stands for Happy Planet Index) are by default mapped with a class interval scale to the polygon fill. This can be set explicitly with `tm_scale_intervals`, via which the intervals can be specified, as well as the visual values (in this case polygon fill colors):

```{r}
s + tm_polygons(
  fill = "HPI",
  fill.scale = tm_scale_intervals(
    style = "fisher",      # a method to specify the classes
    n = 7,                 # number of classes
    midpoint = 38,         # data value mapped to the middle palette color
    values = "pu_gn_div"   # color palette; 
                           # run cols4all::c4a_gui() to explore color palettes
  ))
```

The `style` parameter within `tm_scale_intervals` has a variety of options, including:

- `fixed`
- `sd`
- `equal`
- `pretty`
- `quantile`
- `fisher`
- `jenks`

To specify the classification breaks manually, use `style = "fixed"` and specify the breaks using `breaks = c(0,10,20,30,40,50,60)`:

```{r}
s + tm_polygons(
  fill = "HPI",
  fill.scale = tm_scale_intervals(
    n = 6, # for n classes 
  	style = "fixed",    
    breaks = c(0,10,20,30,40,50,60), # you need n+1 number of breaks
    values = "pu_gn_div"
  ))
```

By default the legend will show bins. Alternatively, the breaks can be printed between the colors:

```{r}
s + tm_polygons(
  fill = "HPI",
  fill.scale = tm_scale_intervals(
    n = 6,
  	style = "fixed",    
    breaks = c(0,10,20,30,40,50,60),
    values = "pu_gn_div",
    label.style = "continuous"
  ))
```


## Scales: numeric data (continuous)

An alternative for numeric data variables is the continuous scale:

```{r, fig.height = 3.5}
s + 
  tm_polygons(
    fill = "HPI",
    fill.scale = tm_scale_continuous(
      limits = c(10, 60),
      values = "scico.hawaii"))
```

By default, the limit (so the value range that is mapped to a color gradient) is taken from the actual data value range `range(World$HPI, na.rm = T)`), which is `r range(World$HPI, na.rm = T)`. In this example rounded numbers 10 and 60 are specified here. If needed, the position of ticks can be specified via the argument `ticks`.


## Scales: categorical data

```{r, fig.height = 3.5}
s + 
  tm_polygons(
    fill = "economy",
    fill.scale = tm_scale_categorical())
```

```{r, fig.height = 3.5}
s + 
  tm_polygons(
    fill = "income_grp",
    fill.scale = tm_scale_ordinal(values = "matplotlib.summer"))
```

