---
title: "tmap basics: facets"
output: 
  bookdown::html_vignette2:
pkgdown:
  as_is: true
template:
  math-rendering: mathjax
bibliography: '`r system.file("tmap.bib", package="tmap")`'
csl: "`r system.file('ieee.csl', package = 'tmap')`"
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  out.width = "100%",
  dpi = 300,
  fig.width = 7.2916667,
  comment = "#>"
)
hook_output <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

```


```{r, echo = FALSE, message = FALSE}
library(tmap)
tmap_options(scale = 0.75)
```



### About the data

Three spatial data objects of the Netherlands are contained in tmap: `NLD_prov`, `NLD_muni`, and `NLD_dist`, at respectively province, municipality and district level. The columns are the following data variables plus an additional geometry column which contains the geometries (see sf package):

```{r}
names(NLD_muni)
```


## Facets: one dimensional

Facets or *small multiples* are small visualizations (maps), one for each group. 
The groups can be defined using a categorical data variable. In the following plot, a facet is created for each province (a data variable in `NLD_muni`):

```{r, fig.height = 5}
tm_shape(NLD_muni) +
  tm_polygons("employment_rate") +
tm_facets("province")
```

In this example, there is *one facet variable*. For these kind of plots, and the facets are either **wrapped** or **stacked**, with `tm_facets_wrap()` or `tm_facets_stack()`. The general purpose function `tm_facets()` determines the faceting type, which is in this example `wrap`.

Groups can also be defined by specifying multiple data variable names for one visual variable:

```{r, fig.height = 5}
tm_shape(NLD_muni) +
  tm_polygons(
  	fill = c("employment_rate", "dwelling_value", "income_high"),
  	fill.scale = 
  	  list(tm_scale_intervals(values = "-brewer.reds"),
  	  	 tm_scale_intervals(values = "brewer.purples"),
  	  	 tm_scale_intervals(values = "brewer.blues")))
```

Because different data variables are used, tmap uses (by default) multiple scales, one for each variable. These can be changed by putting the `tm_scale()` functions in a list, as shown above. In case the same scale should be used, `fill.free` should be set to `FALSE`:

```{r, fig.height = 5}
tm_shape(NLD_muni) +
  tm_polygons(
  	fill = c("pop_0_14", "pop_25_44", "pop_65plus"),
  	fill.legend = tm_legend("Percentage"),
  	fill.free = FALSE)
```


## Facets: two or three dimensions

Facets can also be defined for two or three dimensions, obtained via `tm_facets_grid()` These three dimensions correspond to *rows*, *columns*, and *pages* respectively. The most common use case is two just dimensions, rows and columns.
In that case, `tm_facets_grid()` is used.


```{r, fig.height = 5}
tm_shape(NLD_muni) +
  tm_polygons(
  	fill = c("employment_rate", "dwelling_value", "income_high"),
  	fill.scale = 
  	  list(tm_scale_intervals(values = "-brewer.reds"),
  	  	 tm_scale_intervals(values = "brewer.purples"),
  	  	 tm_scale_intervals(values = "brewer.blues"))) +
tm_facets_grid(columns = "province")
```

## Facets as 'pages' (use case: animations)

```{r, eval=FALSE}
(tm = tm_shape(NLD_muni) +
  tm_polygons(
  	fill = c("pop_0_14", "pop_15_24", "pop_25_44", "pop_45_64", "pop_65plus"),
  	fill.legend = tm_legend("Percentage"),
  	fill.free = FALSE) +
tm_facets_pagewise())
```

will produce 5 plots, one for each age class (defined using the `fill` argument).

This plot can be exported as animation via `tmap_animation`.

```{r, eval=FALSE}
tmap_animation(tm, filename = "NLD_pop_age_class.gif")
```

