---
title: "tmap basics: modes"
output: 
  bookdown::html_vignette2:
bibliography: '`r system.file("tmap.bib", package="tmap")`'
csl: "`r system.file('ieee.csl', package = 'tmap')`"
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 8,
  comment = "#>"
)
hook_output <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

```


```{r, echo = FALSE, message = FALSE}
library(tmap)
tmap_options(scale = 0.75)
```



## Modes

tmap facilitates two output modes: 

* `"plot"`, which produces static maps, and 
* `"view"` which produces (using the same tmap code) interactive maps (using the JavaScript library Leaflet as backend).

Via the extension package, `tmap.mapgl`, two interactive new modes are available, `"mapbox"` and `"maplibre"`, as demonstrated below. 

 
## Switching between modes

```{r}
## current mode
tmap_mode()

## to view mode
tmap_mode("view")

## back to plot mode
tmap_mode("plot")
```

The handy function `ttm()` is used to toggle between the modes. When more than two modes are loaded, `rtm()` can be used to rotate between modes.


## Mode `"plot"`

We start with creating the plot and assign it to a variable called `tm`.

```{r}
tm = tm_shape(World, crs = 8857) +
  tm_polygons(
    fill = "press",
    fill.scale = tm_scale_intervals(values = "pu_gn")) +
tm_shape(metro) +
  tm_bubbles(
    size = "pop2020",
    fill = "gold",
    size.scale = tm_scale_continuous(values.scale = 0.8, n = 8))
```

We are in `"plot"` mode. Now we can plot the map by printing the `tm` object:

```{r, fig.height = 3.5, out.width = "100%", dpi = 300, fig.width = 7.2916667}
tm
```


## Mode `"view"`

The same map in view mode:

```{r}
ttm()
tm
```

Note that there is a big difference: in `"view"` mode there are basemaps, and in `"plot"` mode none. This is caused by different default options.

Basemaps can be enabled or disabled via `tm_basemap()`: 

```{r, fig.height = 3.5}
tm + tm_basemap(NULL)
```

See [vignette about basemaps](https://r-tmap.github.io/tmap/articles/basics_basemaps).

## Mode specific options

Mode specific layout options can be set via `tm_plot()` and `tm_view()`. The number of options in `tm_plot()` is limited to just two, because it uses all general purpose options. In contrast, `tm_view()` contains more options, e.g. the position of the control box and the default zoom level:

```{r, fig.height = 3.5, eval=FALSE}
tm +
	tm_view(control.position = c("left", "bottom"),
			set_view = 2)
```

For a more detailed description of the available options, see the [vignette about options](https://r-tmap.github.io/tmap/articles/adv_options).

## Modes `"mapbox"` and `"maplibre"`.

The new package `tmap.mapgl` offers two new modes. The first is `"mapbox"` for which an API key is required, which is free for personal use. The second is `maplibre` for which no API key is required. `maplibre` is a fork of `mapbox` before it sent from open to closed source.

These modes are not as feature rich (yet) as the `"view"` mode, but they do offer a lot of new features: globe view and 3d polygons. This is possible because of the support of WebGL, a fast rendering technique. Therefore, `maplibre` (because it is open source) is [considered the way forward in web mapping](https://tim-salabim.codeberg.page/sotwm/sotwm.html).

```{r, fig.height = 3.5}
library(tmap.mapgl)
```

When more than two modes are loaded (now four), `rtm()` can be used to rotate between modes:

```{r}
rtm()
rtm()
```

This (and `ttm()`) is especially useful in interactive sessinos. For scripts, it is recommend to use `tmap_mode()` explicitly.

```{r, fig.height = 3.5}
tmap_mode("maplibre")
tm
```



