---
title: "tmap basics: components"
output: 
  bookdown::html_vignette2:
pkgdown:
  as_is: true
template:
  math-rendering: mathjax
bibliography: '`r system.file("tmap.bib", package="tmap")`'
csl: "`r system.file('ieee.csl', package = 'tmap')`"
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  out.width = "100%",
  dpi = 300,
  fig.width = 7.2916667,
  comment = "#>"
)
hook_output <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

```


```{r, echo = FALSE, message = FALSE}
library(tmap)
tmap_options(scale = 0.75)
```

## What are 'components' in tmap?

Components are non-spatial parts of the plots:

* titles and credits: [tm_title()] and [tm_credits()]
* scale bars and compasses: [tm_scalebar()] and [tm_compass()]
* legends: specified with the `.legend` compoments in the data layer functions (e.g. [tm_polygons()]).

## Title and credits

`tm_title()` and `tm_credits()` add text to the map in the form a title and a credits/attribution text respectively. The underlying function and all the options are the same, but the default value differ: titles are by default larger and placed above the map (outside the frame), while credits are place inside the frame at the bottom right. 


```{r, fig.height = 6}
tm_shape(NLD_muni) +
  tm_polygons(
    fill = "edu_appl_sci") +
tm_title("Population share with (applied) university degree in 2022") +
tm_credits("Statistics Netherlands (CBS)")
```


##  Scale bar and compass

```{r, fig.height = 6}
tm_shape(NLD_muni) +
  tm_polygons() +
	tm_compass(type = "8star") +
	tm_scalebar()
```

## Positioning of components (basics)

The position of a map component can be specified with the general argument `position`.

It takes a vector of two, the first is the horizontal position (`"left"`, `"center"`, `"right"`), the second one the vertical position (`"top"`, `"center"`, `"bottom"`)

```{r, fig.height = 6}
tm_shape(NLD_muni) +
  tm_polygons(
    fill = "edu_appl_sci",
    fill.legend = tm_legend(position = c("left", "top"))) +
tm_credits("Statistics Netherlands (CBS)", position = c("left", "bottom")) +
tm_compass(type = "8star", position = c("right", "bottom")) +
tm_scalebar(position = c("right", "bottom"))
```

If the components should be placed tighter to the frame, use capital case instead; e.g. `c("LEFT", "BOTTOM") in stead of `c("left", "bottom")`.

There are much more options to position map componets via `tm_pos()`, see the [vignette about positions](https://r-tmap.github.io/tmap/articles/41_position)
