---
title: "tmap basics: basemaps"
output: 
  bookdown::html_vignette2:
pkgdown:
  as_is: true
template:
  math-rendering: mathjax
bibliography: '`r system.file("tmap.bib", package="tmap")`'
csl: "`r system.file('ieee.csl', package = 'tmap')`"
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  out.width = "100%",
  dpi = 300,
  fig.width = 7.2916667,
  comment = "#>"
)
hook_output <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

```


```{r, echo = FALSE, message = FALSE}
library(tmap)
tmap_options(scale = 0.75)
```



## Basemaps

By default, basemaps are only shown in `"view"` mode, so not in `"plot"` mode. Basemaps can be enabled or disabled via `tm_basemap()`. 

## Plot mode

To enable a basemap in plot mode, the package [`maptiles`](https://github.com/riatelab/maptiles/) is required. When specified without arguments, the default basemap server is "Esri.WorldGrayCanvas" (run `tmap_options("basemap.server")` to see this option).

```{r, fig.height = 3.5}
tm_shape(metro) +
	tm_bubbles(size = "pop2020") +
	tm_basemap()
```

This can be changed. The options are:

```{r}
names(maptiles::get_providers())
```

See [previews](https://leaflet-extras.github.io/leaflet-providers/preview/)

```{r, fig.height = 3.5}
tm_shape(metro) +
	tm_bubbles(size = "pop2020") +
	tm_basemap("Esri.OceanBasemap")
```

## View mode


```{r}
tmap_mode("view")
tm_shape(metro) +
	tm_bubbles(size = "pop2020") +
	tm_basemap("Esri.OceanBasemap")
```


Disabling basemaps:

```{r}
tmap_mode("view")
tm_shape(World) +
  tm_polygons(
    fill = "grey80", 
    col = "grey60") +
tm_shape(metro) +
	tm_bubbles(size = "pop2020") +
	tm_basemap(NULL)
```

Now it is also possible to use different map projections (see [vignette](https://r-tmap.github.io/tmap/articles/12_foundations_crs)):

```{r}
tmap_mode("view")
tm_shape(World, crs = "+proj=robin") +
  tm_polygons(
    fill = "grey80", 
    col = "grey60") +
tm_shape(metro) +
	tm_bubbles(size = "pop2020") +
	tm_basemap(NULL)
```

