---
title: "tmap basics: visual variables"
output: 
  bookdown::html_vignette2:
pkgdown:
  as_is: true
template:
  math-rendering: mathjax
bibliography: '`r system.file("tmap.bib", package="tmap")`'
csl: "`r system.file('ieee.csl', package = 'tmap')`"
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  out.width = "100%",
  dpi = 300,
  fig.width = 7.2916667,
  comment = "#>"
)
hook_output <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

```


```{r setup, echo = FALSE, message = FALSE}
library(tmap)
tmap_options(scale = 0.75)
```

## Introduction

**tmap** is an R package for spatial data visualization.

### Map layers

A (thematic) map consists of one or more map layers. Each map layer has a specific set of variables that determine how the objects of that layer are drawn. A **visual variable**
changes the appearance of a spatial object, e.g. fill color or line width.

The following table shows which visual variables are used in standard map layers.

|Map layer		|Visual variables				|
|:--------------|:--------------------------------------------------------------|
|`tm_polygons`	|`fill` (fill color), `fill_alpha` (fill transparency), <br> `col` (border color), `col_alpha` (border color transparency) <br> `lwd` (border line width), `lty` (border line type)	|
|`tm_symbols` |`size`, `shape` <br> `fill` (fill color), `fill_alpha` (fill transparency), <br> `col` (border color), `col_alpha` (border color transparency) <br> `lwd` (border line width) `lty` (border line type) |
|`tm_lines`		|`col` (border color), `col_alpha` (border color transparency) <br> `lwd` (line width) `lty` (line type) |
|`tm_raster`	|`col` (color),	`col_alpha` (transparency)							|
|`tm_text`	|`text` (the text itself), `size` (font size), <br> `col` (border color), `col_alpha` (border color transparency) <br> `fontface` (font face), `fontfamily` (font family) |

## About the example data

A spatial data object contained in **tmap** is called `World`. It is a data frame with a row for each country. The columns are the following data variables plus an additional geometry column which contains the geometries (see sf package):

```{r names}
names(World)
```

We specify this object with `tm_shape`
<!-- (see other vignette)-->
and for convenience assign it to `s`: 

```{r shape}
s <- tm_shape(World, crs = "+proj=eqearth")

```

## Constant visual values

Each visual variable can be assigned a constant value. E.g. `fill = "red"` or `size = 2`.

An example where several visual variables are specified with constant values:

```{r constant, fig.height = 3.5}
s + 
  tm_polygons(
    fill = "#ffce00", # fill color
    col = "black",    # line color
    lwd = 0.5,        # line width
    lty = "dashed")   # line type
```

For advanced users: the default constant values are specified for combinations of visual variables and layer type. See `tmap_options("value.const")`



## Data-driven visual variables

Each visual variable argument can also be specified with a data variable (e.g., a column name):

```{r polygons}
s + tm_polygons(fill = "press")    # a column in World
```

Another example, where a data-driven symbol layer is drawn on top of a polygon layer

```{r symbols}
s + 
  tm_polygons(fill = "grey90") +   # constant fill color 
  tm_symbols(size = "pop_est",     # data variable, mapped to symbol size
  		     fill = "well_being",  # data variable, mapped to symbol fill color
  		     shape = "income_grp") # data variable, mapped to symbol shape

```

In the following example a data-driven text layer is plotted.

```{r text}
s + 
  tm_polygons(fill = "economy") +
  tm_text(text = "name", size = "area")
```

### Scales and legend

For each data-driven visual variable, a scale and legend can be specified. See other vignettes.

## Facets

A facet map is created by specifying two data variables (columns in `World`) to the visual variable `fill`:

```{r facet, fig.height = 3.5}
s + tm_polygons(fill = c("well_being", "life_exp")) 
```

Improve the titles and remove (duplicate) legend titles

```{r, fig.height = 7}
s + 
  tm_polygons(
    fill = c("well_being", "life_exp"),
    fill.legend = tm_legend("")) +
tm_layout(panel.labels = c("Well Being", "Life Expectancy"))
```
