---
title: "tmap basics: inset maps and minimaps"
output: 
  bookdown::html_vignette2:
pkgdown:
  as_is: true
template:
  math-rendering: mathjax
bibliography: '`r system.file("tmap.bib", package="tmap")`'
csl: "`r system.file('ieee.csl', package = 'tmap')`"
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  out.width = "100%",
  dpi = 300,
  fig.width = 7.2916667,
  comment = "#>"
)
hook_output <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

```


```{r, echo = FALSE, message = FALSE}
library(tmap)
tmap_options(scale = 0.75)
```

## Introduction

An inset map or minimap is a small map embedded in the main map.

* We use `tm_inset()` to zoom in on areas that are too cluttered at the main maps's scale.
* We use `tm_minimap(()` to show a broader geographic context; where on earth is the main map located?

Note that both `tm_inset()` and `tm_minimap()` are map components (see [vignette](https://r-tmap.github.io/tmap/articles/basics_components))

## Minimaps


```{r, fig.height = 6}
tmap_mode("plot")
tm_shape(NLD_muni) +
	tm_polygons(
		fill = "edu_appl_sci") +
	tm_title("Population share with (applied) university degree in 2022") +
	tm_credits("Statistics Netherlands (CBS)") + 
	tm_minimap(position = c("left", "top"))
```

The implementation in plot mode is experimental (and new as of tmap 4.2). It features a small globe. There are no options implemented yet to change the appearance.
Note that it takes a few seconds to run due to the applied orthographic [map projection](https://r-tmap.github.io/tmap/articles/foundations_crs).

The minimap is interactive in view mode.

```{r, fig.height = 6}
tmap_mode("view")
tmap_last()
```

Note that the minimap is also supported in two other modes, `"maplibre"` and `"mapbox"` via the extension package [**tmap.mapgl**](https://r-tmap.github.io/tmap.mapgl/).


## Inset maps

```{r, fig.height = 6}
tmap_mode("plot")
tm_shape(NLD_dist) +
	tm_polygons(
		fill = "edu_appl_sci",
		col = NULL) +
tm_shape(NLD_muni) +
	tm_borders() +
	tm_title("Population share with (applied) university degree in 2022") +
	tm_credits("Statistics Netherlands (CBS)") + 
	tm_title("Amsterdam", size = 0.8, position = c("left", "top")) +
	tm_inset(bb("Amsterdam"), position = c("left", "top")) +
	tm_title("The Hague", size = 0.8, position = c("left", "top")) +
	tm_inset(bb("The Hague"), position = c("left", "top"))
```

Currently, inset maps are only implemented in `"plot"` mode.
