---
title: "tmap example: topographic map (Africa)"
output: 
  bookdown::html_vignette2:
pkgdown:
  as_is: true
template:
  math-rendering: mathjax
bibliography: '`r system.file("tmap.bib", package="tmap")`'
csl: "`r system.file('ieee.csl', package = 'tmap')`"
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  out.width = "100%",
  dpi = 300,
  fig.width = 7.2916667,
  comment = "#>"
)
hook_output <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

```

```{r, message = FALSE}
library(tmap)
library(dplyr)
library(sf)
tmap_options(scale = 0.75)
```


## About the data

We use several spatial data object that are contained in tmap. The first is called `World`, which we only use for country borders. We'll only select the African countries:

```{r}
metroAfrica = sf::st_intersection(metro, World[World$continent == "Africa", ])
Africa = World[World$continent == "Africa", ]
```

The second is `land` which is a `stars` object that contains land cover. The third object is `metro` an `sf` object with metropolitan areas. The fourth and final ovject is `World_rivers`.

## Topographic map

```{r, fig.height = 6}
tm_shape(land) +
  tm_raster("cover_cls",
    col.legend = tm_legend("Land use")) +
tm_shape(World_rivers) +
  tm_lines(
  	lwd = "strokelwd", 
  	lwd.scale = tm_scale_asis(values.scale = .5),
    col = "#A6CEE3") +
tm_shape(Africa, is.main = TRUE) +
  tm_borders() +
tm_shape(metroAfrica) +
  tm_symbols(fill = "red", shape = "pop2020", size = "pop2020",
    size.scale = tm_scale_intervals(
      breaks = c(1, 2, 5, 10, 15, 20, 25) * 1e6,
      values.range = c(0.2,2)
    ),
    size.legend = tm_legend("Population in 2020"),
    shape.scale = tm_scale_intervals(
      breaks = c(1, 2, 5, 10, 15, 20, 25) * 1e6,
      values = c(21, 23, 22, 21, 23, 22)
    ),
    shape.legend = tm_legend_combine("size")) +
  tm_labels("name") +
tm_credits("United Nations, Department of Economic and Social Affairs, Population Division (2014). World Urbanization Prospects.\nProduction of Global Land Cover Data - GLCNMO2008.", position = tm_pos_out("center", "bottom"))
```

## Breaking down in layers

If this code chunk is overwhelming, let's that a look at the layer groups:

### Land use layer

```{r, fig.height = 3.5}
tm = 
tm_shape(land) +
  tm_raster("cover_cls",
    col.legend = tm_legend("Land use"))
tm
```


### Adding rivers

```{r, fig.height = 3.5}
tm = tm + 
tm_shape(World_rivers) +
  tm_lines(
  	lwd = "strokelwd", 
  	lwd.scale = tm_scale_asis(values.scale = .5),
    col = "#A6CEE3")
tm
```

### African countries

Note that the first two layers are from a global data set (so not just Africa). 
Note that in the final plot this raster is cropped to the bounding box of Africa.
We'll use `is.main` for the the spatial object `Africa` to do the actual cropping.

```{r, fig.height = 6}
tm = tm +
tm_shape(Africa, is.main = TRUE) +
  tm_borders()
tm
```

### African cities

```{r, fig.height = 6}
tm = tm +
tm_shape(metroAfrica) +
  tm_symbols(fill = "red", shape = "pop2020", size = "pop2020",
    size.scale = tm_scale_intervals(
      breaks = c(1, 2, 5, 10, 15, 20, 25) * 1e6,
      values.range = c(0.2,2)
    ),
    size.legend = tm_legend("Population in 2020"),
    shape.scale = tm_scale_intervals(
      breaks = c(1, 2, 5, 10, 15, 20, 25) * 1e6,
      values = c(21, 23, 22, 21, 23, 22)
    ),
    shape.legend = tm_legend_combine("size")) +
  tm_labels("name")
tm
```

See the [vigette about scale](https://r-tmap.github.io/tmap/articles/basics_scales) and [about legends](https://r-tmap.github.io/tmap/articles/basics_legends).
