---
title: "tmap extensions: tmap.cartogram"
output: 
  bookdown::html_vignette2:
bibliography: '`r system.file("tmap.bib", package="tmap")`'
csl: "`r system.file('ieee.csl', package = 'tmap')`"
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 8,
  comment = "#>"
)
hook_output <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

```


```{r, echo = FALSE, message = FALSE}
library(tmap)
library(tmap.cartogram)
tmap_options(scale = 0.75)
```

## Cartograms

In cartograms, areal regions are distorted such that the obtained area sizes are proportional to a quantitative variable. There are three types: contiguous, non-contiguous and dorling cartogram.

The extension package [**tmap.cartogram**](https://r-tmap.github.io/tmap.cartogram/) provides new layer functions for tmap to create various types of cartograms.

It is strongly recommended to use an equal area crs (see [instructions](https://r-tmap.github.io/tmap.cartogram/articles/intro#map-projection)).

## Contiguous cartogram

The shapes of the polygons are distorted, where the neighborhood relationships between them are preserved as much as possible.

```{r, fig.height=6}
Africa = World[World$continent == "Africa", ]

tm_shape(Africa, crs = "+proj=robin") +
  tm_cartogram(
    size = "*pop_est", 
	fill = "life_exp",
	fill.scale = tm_scale_intervals(values = "-cols4all.bu_br_div"),
	fill.legend = tm_legend("Age"),
	options = opt_tm_cartogram(itermax = 15)) +
tm_title("Life Expectancy") +
tm_animate_fast(play = "pingpong")
```

## Non-contiguous cartograms

Here, the polygons are only resized while they keep their shape.

```{r, fig.height=6}
tm_shape(Africa, crs = "+proj=robin") +
  tm_cartogram_ncont(
  	size = "pop_est", 
	fill = "inequality",
	fill.scale = tm_scale_continuous(values = "cols4all.pu_gn_div", values.range = c(0, 0.5)),
	fill.legend = tm_legend(""),
	options = opt_tm_cartogram_ncont()) +
  tm_text("name", options = opt_tm_text(point.label = TRUE)) +
tm_title("Income inequality (Gini coefficient)")
```

## Dorling cartogram

Polygons are replaced by non-overlapping bubbles.

```{r, fig.height=4}
tm_shape(World, crs = "+proj=robin") +
  tm_cartogram_dorling(
    size = "pop_est", 
	fill = "press",
	fill.scale = tm_scale_continuous(values = "cols4all.pu_gn_div", midpoint = 50),
	fill.legend = tm_legend("", height = 30)) +
tm_title("World Press Freedom Index")
```
