---
title: "tmap: get started!"
output: 
  rmarkdown::html_vignette:
    keep_md: true 
    toc: true
self_contained: no
vignette: >
  %\VignetteIndexEntry{tmap: get started!}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(collapse = T, fig.width=8, fig.height=4)
#devtools::install_github("mtennekes/tmaptools")
library(tmap)
library(sf)
```


With the tmap package, thematic maps can be generated with great flexibility. The syntax for creating plots is similar to that of `ggplot2`, but tailored to maps. This vignette is for those who want to get started with tmap within a couple of minutes. A more detailed description of tmap can be found in the [JSS paper](https://www.jstatsoft.org/article/view/v084i06). Note that this paper is written for tmap version 1.11-2. The changes in version 2.0 are described in [`vignette("tmap-changes")`](../doc/tmap-changes-v2.html)


### Hello World!

```{r}
data("World")

tm_shape(World) +
	tm_polygons("HPI")
```

The object `World` is a spatial `sf` object; it is a `data.frame` with a special column that contains the geometry of the units, in this case polygons. In order to plot it in tmap, you first need to specify it with `tm_shape`. With the `+` operator plotting layers can be added, in this case `tm_polyongs`. There are many layer functions in tmap, which can easlily be found in the documentation by their `tm_` prefix.

### Interactive maps

Each map can be plotted as a static image and interactively viewed. These are called respectively the \code{"plot"} and the \code{"view"} mode respectively. The mode can be set with the function `tmap_mode`, and toggling between the modes can be done with `ttm()`.


```{r}
tmap_mode("view")

tm_shape(World) +
	tm_polygons("HPI")
```


### Multiple shapes and layers

A shape is an spatial object (with a class from `sf`, `sp` or `raster`). Mutliple shapes can be plotted, and also multiple layers per shape:

```{r}
data(World, metro, rivers, land)

tmap_mode("plot")
tm_shape(land) +
	tm_raster("elevation", palette = terrain.colors(10)) +
tm_shape(World) +
	tm_borders("white", lwd = .5) +
	tm_text("iso_a3", size = "AREA") +
tm_shape(metro) +
	tm_symbols(col = "red", size = "pop2020", scale = .5) +
tm_legend(show = FALSE)
```

### Facets

```{r}
tmap_mode("view")
tm_shape(World) +
	tm_polygons(c("HPI", "economy")) +
	tm_facets(sync = TRUE)
```

```{r}
tmap_mode("plot")

data(NLD_muni)

NLD_muni$perc_men <- NLD_muni$pop_men / NLD_muni$population * 100

tm_shape(NLD_muni) +
	tm_polygons("perc_men", palette = "RdYlBu") +
	tm_facets("province")
```

### Basemaps and overlay tile maps

Tiled basemaps can be added with the layer function `tm_basemap`. Semi-transparent overlay maps (for example annotation labels) can be added with `tm_tiles`.

```{r}
tm_basemap("Stamen.Watercolor") +
tm_shape(metro) + tm_bubbles(size = "pop2020", col = "red") +
tm_tiles("Stamen.TonerLabels")
```

See a [preview of the available tilemaps](http://leaflet-extras.github.io/leaflet-providers/preview/). This list is also accessible in R: `leaflet::providers`.

### Options and styles

The functions `tm_layout` and `tm_view` are used to specify the map layout and the interactive aspects respectively. These functions can be used in the same way as the layer functions, e.g.

```{r}
tmap_mode("plot")

tm_shape(World) +
	tm_polygons("HPI") +
	tm_layout(bg.color = "skyblue", inner.margins = c(0, .02, .02, .02))
```

However, these options, as well as a couple of other, can also be set within an R session with `tmap_options`, which works in the same way as the base R function `options`:

```{r}
tmap_options(bg.color = "black", legend.text.color = "white")

tm_shape(World) +
	tm_polygons("HPI", legend.title = "Happy Planet Index")
```

A style is a certain configuration of the tmap options. 

```{r}
tmap_style("classic")

tm_shape(World) +
	tm_polygons("HPI", legend.title = "Happy Planet Index")
```

```{r}

# see what options have been changed
tmap_options_diff()

# reset the options to the default values
tmap_options_reset()
```

### Exporting maps


```{r}
tm <- tm_shape(World) +
	tm_polygons("HPI", legend.title = "Happy Planet Index")

## save an image ("plot" mode)
tmap_save(tm, filename = "world_map.png")

## save as stand-alone HTML file ("view" mode)
tmap_save(tm, filename = "world_map.html")

```

### leaflet and shiny integration

Maps made with tmap can be exported to leaflet with the function `tmap_leaflet`. Hence, it is possible to make use of leaflet own functionalities. This function is also needed when using tmap in shiny:


```{r, eval = FALSE}
# in ui part:
leafletOutput("my_tmap")

# in server part
output$my_tmap = renderLeaflet({
	tm <- tm_shape(World) + tm_polygons("HPI", legend.title = "Happy Planet Index")
    tmap_leaflet(tm)
})
```

See full [tmap - shiny example](https://github.com/mtennekes/tmap/blob/master/demo/tutorials/shiny_tmap.R)


### Quick thematic map

Maps can also be made with one function call. This function is `qtm`:

```{r}
qtm(World, fill = "HPI", fill.pallete = "RdYlGn")
```



<!-- ### Tips n' tricks -->


<!-- 1. A manual legend can be generated [`tm_add_legend`](../html/tm_add_legend.html): -->
<!-- ```{r, fig.height=3} -->
<!-- data(World) -->

<!-- tm_shape(World) + -->
<!-- 	tm_fill() + -->
<!-- tm_shape(rivers) + -->
<!-- 	tm_lines(col="dodgerblue3") + -->
<!-- 	tm_add_legend(type="line", col="dodgerblue3", labels = "Rivers", title="World map") + -->
<!-- tm_format("World") -->
<!-- ``` -->

<!-- 2. Each drawing element has a scalar argument called `scale`. The overall scaling and font sizes can be set by the `scale` argument in `tm_layout`. -->

<!-- 3. Arguments of the bounding box function [`bb`](../html/bb.html) can be passed directly to [`tm_shape`](../html/tm_shape.html): -->
<!-- ```{r, fig.height=5} -->
<!-- tm_shape(World, bbox = "India") + -->
<!-- 	tm_polygons("MAP_COLORS", palette="Pastel2") + -->
<!-- tm_shape(metro) + -->
<!-- 	tm_bubbles("pop2010", title.size = "Population") + -->
<!-- 	tm_text("name", size = "pop2010", legend.size.show = FALSE, root=8, size.lowerbound = .7, auto.placement = TRUE) -->
<!-- ``` -->

