---
title: "geo in a nutshell"
author: "Martijn Tennekes"
output: html_document
---


```{r setup, include=FALSE}
library(knitr)
devtools::load_all("../")
```

### Introduction

With the geo package, statistical thematic maps can be generated with great flexability. The syntax for creating plots is very similar to `ggplot2`. It also contains functions to deal with shape objecte, where many are convenient wrappers of functions from other packages such as `sp` and `rgeos`.


### Shape objects

We refer to **shape objects** as objects from the class `Spatial` from the package `sp`. The six supported subclasses are:

|   	| Without data	| With data   	|
|---	|---	|---	|
|Polygons   	| SpatialPolygons  	| SpatialPolygonsDataFrame  	|
|Points   	| SpatialPoints  	| SpatialPointsDataFrame  	|
|Lines   	| SpatialLines  	| SpatialLinesDataFrame  	|

Obviously, shape objects with data (the right-hand side column) are recommended, since data is what we want to show.

Load shape object of Europe (contained in this package):
```{r}
data(Europe)
```

Shape objects in ESRI format can be read with `read_shape` and saved with `write_shape`.

Projection can be get and set with `get_projection` and `set_projection` respectively.


### Quick plots

The `geo` function is `geo`'s equivalent to `ggplot2`'s `qplot`.

```{r}
geo(Europe)
```

A choropleth is created with one line of code:

```{r}
geo(Europe, fill="gdp_cap_est", text="iso_a3", text.cex="pop_est", title="GDP per capita", textNA="Non-EU countries")
```





